---
title: "Discrete outcome & discrete predictor assumptions and interpretation"
subtitle: "Sometimes it is just black and white"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**CONTENTS**"
    output-file: "2.2-lnOR-assumptions.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
---

```{r}
#| eval: true
#| echo: false
#| out-width: "70%"

rm(list = ls())
library("metaDigitise")
library("ggplot2")
library("data.table")
library("ggpattern")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

# Source our custom functions:
source("scripts/functions/eff_size.R")

```

The log odds ratio (lnOR) and risk ratio (lnRR) effect sizes are based on two-group contingency tables. These study designs use a discrete treatment and have a discrete binary outcome. Are they dead or alive, did they breed or not, etc. Such data looks like this, where a, b, c, d are integer count values.

|         | Alive | Dead | Total      |
|---------|-------|------|------------|
| Drug    | $a$   | $b$  | $n1 = a+b$ |
| Placebo | $c$   | $d$  | $n2 = c+d$ |

These outcomes are binomial, i.e., are discrete processes with binary outcomes that occur to individuals. Individuals can be both individual organisms as well as individual sites/islands/plots, i.e., in terms of presence or absence of some binary condition.

Tertiary outcomes binomial outcomes are also possible and need to be handled with alternative effect sizes. For example, the frequency of alleles (AA, Aa, aa) in a population can be described with the Hardy-Weinburg Equilibrium effect size (REF). These are not discussed in this tutorial. 

## log Odds Ratio (lnOR) {#sec-lnOR}

The log odds ratio is the ratio of two odds:

$$lnOR = log(\frac{a * d}{b * c})$$

### Assumptions {#sec-lnOR-assumptions}

Blah blah blah

### Interpretation {#sec-lnOR-interpretation}

To give an example, imagine a study reported on survivorship outcomes of rodents based on a drug versus of a placebo, as follows.

|         | Alive | Dead |
|---------|-------|------|
| Drug    | 35    | 12   |
| Placebo | 15    | 21   |

lnOR can be calculated by plugging these values into escalc:

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

escalc(measure = "OR",
       ai = 35, bi = 12,
       ci = 15, di = 21)

```

To interpret this log odds ratio ($lnOR$), you should first convert it to an odds ratio ($OR$):

$$OR = e^{lnOR}$$

```{r}
exp(1.4069) |> round()
```

An odds ratio of 4 indicates that the odds of a rodent dying are 4 times as high when given the placebo instead of the drug. An odds ratio of 1 indicates that the odds of mortality are identical between the two treatments.

## log Risk Ratio (lnRR) {#sec-lnRR}

The risk ratio (or relative risk, lnRR) is similar to lnOR. While lnOR describes the odds of an event occurring, lnRR measures the probability of an event happening in one group relative to the others. The log risk ratio may be more easy to interpret but has numeric properties that make lnOR statistically preferable in many cases [@borenstein2021].

lnRR is defined as:

$$\begin{aligned}
&p1 = a/n1 \\
&p2 = c/n2 \\
&lnRR = log(p1/p2) \\
\end{aligned}$$

We can do something similar with the log risk ratio. Again, we will exponentiate it to convert it from log scale.

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

out <- escalc(measure = "RR",
       ai = 35, bi = 12,
       ci = 15, di = 21)

exp(out$yi[1])

```


**Kyle, can you help with this?**

A risk ratio of 1.79 indicates that the risk of mortality in the placebo group is 1.79 times greater than for the rodents that received the drug.

**Not sure I am correctly interpreting these**

## Continuity corrections {#sec-continuity-corrections}

lnOR breaks down whenever $a$, $b$, $c$, or $d$ equal 0. To overcome this, we apply a continuity correction (Yates 1934) by adding 0.5 to all cells if any cell equals 0.

lnRR breaks down when $a$ or $c$ equal 0 and thus also requires a continuity correction. For lnRR, you add 0.5 to the affected group (not both groups) and add 1 to n1 or n2, respectively.


# Alternative versions
There are alternative estimators for similar data that take into account longitudinal pairing (e.g., through time) or matched pairs of subjects. These are broadly referred to as matched pairs contingency table effect sizes by `escalc`.

Extraction for these effect sizes is similar to lnOR and lnRR except that they take an additional parameter: $rho$, or the correlation between subjects---which is likely unreported and needs to be guessed. Unlike paired versions of SMD/lnRoM/Zmu, these do not appear to be comparable to unpaired estimators of lnOR and lnRR, nor are they widely used in ecology and evolution (REF TO THAT NAKAGAWA PAPER). 

**DELETE:**
The measurement arguments for these effect sizes are given [here](5-escalc-cheatsheet.qmd)

## Matched pairs contingency tables {#sec-paired-contigency-tables}

**Are these paired lnOR????? And, if so, can they be compared to regular lnOR??**

Directly from `escalc`:

"The data for a study examining change in a dichotomous variable gives rise to a paired table, which is of the form:"

|                 | trt 2 outcome 1 | trt 2 outcome 2 |
|-----------------|-----------------|-----------------|
| trt 1 outcome 1 | ai              | bi              |
| trt 1 outcome 2 | ci              | di              |

"Note that ‘trt1’ and ‘trt2’ may be applied to a single group of subjects or to matched pairs of subjects. Also, ‘trt1’ and ‘trt2’ might refer to two different time points (e.g., before and after a treatment). In any case, the data from such a study can be rearranged into a marginal table of the form:"

|                 | outcome 1 | outcome 2 |
|-----------------|-----------|-----------|
| trt 1 or time 1 | ai + bi   | ci + di   |
| trt 2 or time 2 | ai + ci   | bi + di   |

*I'm not sure I follow this*

Options in `escalc`: "MPRR" (marginal log-risk ratio); "MPOR" (marginal log odds ratio); "MPRD" (marginal risk difference)

And for paired: "MPORC" (conditional log odds ratio) and "MRPETO" (conditional log odds ratio estimated with Peto's method)

The paired versions require $r$, just as with paired SMD/lnRoM. In R:

```{r}
# Let's compare MPORC to OR:
escalc(measure = "OR",
       ai = 35, bi = 12,
       ci = 15, di = 21)

escalc(measure = "MPORC",
       ai = 35, bi = 12,
       ci = 15, di = 21,
       ri = 0.5)
# Whoa, nope.


escalc(measure = "OR",
       ai = 35, bi = 12,
       ci = 15, di = 21)

escalc(measure = "MPOR",
       ai = 35, bi = 12,
       ci = 15, di = 21,
       ri = 0.5)
# NOPE

escalc(measure = "RR",
       ai = 35, bi = 12,
       ci = 15, di = 21)

escalc(measure = "MPRR",
       ai = 35, bi = 12,
       ci = 15, di = 21,
       ri = 0.5)
# NOPE
# OK SO DEFINITELY not comparable 

```

