---
title: "Discrete outcome & discrete predictor assumptions and interpretation"
subtitle: "Sometimes it is just black and white"
date: today
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**CONTENTS**"
    output-file: "2.2-lnOR-assumptions.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
csl: ecology.csl
---

```{r}
#| eval: true
#| echo: false
#| out-width: "70%"

rm(list = ls())

if(!requireNamespace("metatools")) {
  pak::pak("ejlundgren/metatools")
}
library("metatools")

library("ggplot2")
library("data.table")
library("ggpattern")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

```

The log odds ratio (lnOR) and risk ratio (lnRR) effect sizes are based on two-group contingency tables. These study designs use a discrete treatment and have a discrete binary outcome. Are they dead or alive, did they breed or not, etc. Such data looks like this, where a, b, c, d are integer count values.

|         | Alive | Dead | Total      |
|---------|-------|------|------------|
| Drug    | $a$   | $b$  | $n1 = a+b$ |
| Placebo | $c$   | $d$  | $n2 = c+d$ |

These outcomes are binomial, i.e., are discrete processes with binary outcomes that occur to individuals. Individuals can be both individual organisms as well as individual sites/islands/plots, i.e., in terms of presence or absence of some binary condition.

Tertiary outcomes binomial outcomes are also possible and need to be handled with alternative effect sizes. For example, the frequency of alleles (AA, Aa, aa) in a population can be described with the Hardy-Weinburg Equilibrium effect size [@Wellek2010-jp]. These are not discussed in this tutorial. 

## log Odds Ratio (lnOR) {#sec-lnOR}

The log odds ratio is the ratio of two odds:

$$
\begin{aligned}
&lnOR = log(\frac{a * d}{b * c}) \\
\end{aligned}
$${#eq-1}

### Assumptions {#sec-lnOR-assumptions}

Blah blah blah

### Interpretation {#sec-lnOR-interpretation}

To give an example, imagine a study reported on survivorship outcomes of rodents based on a drug versus of a placebo, as follows.

|         | Alive | Dead |
|---------|-------|------|
| Drug    | 35    | 12   |
| Placebo | 15    | 21   |

lnOR can be calculated by plugging these values into escalc:

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

escalc(measure = "OR",
       ai = 35, bi = 12,
       ci = 15, di = 21)

```

To interpret this log odds ratio ($lnOR$), you should first convert it to an odds ratio ($OR$):

$$
OR = e^{lnOR}
$${#eq-2}

In `R`:

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"
exp(1.4069) |> round()
```

An odds ratio of 4 indicates that the odds of a rodent dying are 4 times as high when given the placebo instead of the drug. An odds ratio of 1 indicates that the odds of mortality are identical between the two treatments.

## log Risk Ratio (lnRR) {#sec-lnRR}

The risk ratio (or relative risk, lnRR) is similar to lnOR. While lnOR describes the odds of an event occurring, lnRR measures the probability of an event happening in one group relative to the others. The log risk ratio may be more easy to interpret but has numeric properties that make lnOR statistically preferable in many cases [@borenstein2021].

lnRR is defined as:

$$\begin{aligned}
&p1 = a/n1 \\
&p2 = c/n2 \\
&lnRR = log(p1/p2) \\
\end{aligned}$${#eq-3}


### Assumptions
KYLE


### Interpretation
To interpret lnRR, we will also exponentiate it to convert it from log scale:

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

out <- escalc(measure = "RR",
       ai = 35, bi = 12,
       ci = 15, di = 21)

exp(out$yi[1])

```

**Kyle, can you help with this?**

A risk ratio of 1.79 indicates that the risk of mortality in the placebo group is 1.79 times greater than for the rodents that received the drug.

**Not sure I am correctly interpreting these**

# Should I use lnOR or lnRR?

KYLE


## Continuity corrections {#sec-continuity-corrections}

lnOR breaks down whenever $a$, $b$, $c$, or $d$ equal 0. To overcome this, we apply a continuity correction (Yates 1934) by adding 0.5 to all cells if any cell equals 0. 

lnRR breaks down when $a$ or $c$ equal 0 and thus also requires a continuity correction. For lnRR, you add 0.5 to the affected group (not both groups) and add 1 to n1 or n2, respectively.

These continuity corrections are applied by `escalc` by default.

# Alternative versions
There are alternative estimators for similar data that take into account longitudinal pairing (e.g., through time) or matched pairs of subjects. These are broadly referred to as matched pairs contingency table effect sizes by `escalc`.

Extraction for these effect sizes is similar to lnOR and lnRR except that they take an additional parameter: $rho$, or the correlation between subjects---which is likely unreported and needs to be guessed. Unlike paired versions of SMD/lnRoM/Zmu, these do not appear to be comparable to unpaired estimators of lnOR and lnRR, nor are they widely used in ecology and evolution (REF TO THAT NAKAGAWA PAPER). 

The measurement arguments for these effect sizes are given [here](6-escalc-cheatsheet.qmd)
