---
title: "Continuous predictors"
subtitle: "Correlate this"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**CONTENTS**"
    output-file: "2.3-Zr-assumptions.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
---

```{r}
#| eval: true
#| echo: false
#| out-width: "70%"

rm(list = ls())
library("metaDigitise")
library("ggplot2")
library("data.table")
library("ggpattern")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

# Source our custom functions:
source("scripts/functions/eff_size.R")

```

Zr is the dominant effect size for describing correlations between two continuous variables. Zr takes the correlation coefficient and sample size as its inputs. The point estimate itself is the arc-tangent of the correlation coefficient (which must be bounded between -1 and 1). See Appendix (@sec) for sampling variance formulas.

$$\begin{aligned}
&r_{i} \leftarrow \min(1, \max(-1, r_i))\\
&Z_{r_i} = atanh(r_{i})\\
\end{aligned}$$ The sampling variance of Zr depends entirely on sample size and is defined as:

$$Var(Z_r) = \frac{1}{n-3}$$ As you can see, sampling variance is only calculable if $n>3$.

In `escalc`, this looks like:

```{r}
escalc(ri = 0.7, ni = 15,
       measure = "ZCOR")
```

Zr has among the least biased of point estimate and variance formulas.

*What to do about 0/1 or binary outcomes with a continuous predictor?*

## Assumptions

XXXXX

## Interpretation

Zr effect sizes describe the strength and direction of a relationship between two continuous variables.

## Other correlation types

### Point-biserial correlations & biserial correlations

Point-biserial and biserial correlations are oddballs. They are very similar to SMD, in that they are an effect size that summarizes the relationship between a discrete predictor and a continuous outcome variable. However, these effect sizes are a measure of association/correlation and range from -1 to 1, instead of describing the difference in magnitude and direction. Point-biserial correlations are for truly distinct predictor variables while biserial correlations are for continuous variables that have been turned into discrete variables (e.g., large versus small).

In our mind, the biggest advantage of these correlations is that they require fewer assumptions to *convert* to Zr. For example, if you are doing a meta-analysis where the majority of your data are correlations except for one or two studies that use discrete predictors, it may be better to calculate point-biserial or serial correlations for these studies. `escalc` provides a shorthand to directly convert these statistics to Zr:

```{r}
# For the point-biserial correlation coefficient (truly discrete groups), directly to `Zr`
escalc(measure = "ZPB",
      m1i = 13.5, m2i = 10.1,
      sd1i = 1.5, sd2i = 2.1,
      n1i = 15, n2i = 15)


# For the biserial correlation coefficient (for a continuous predictor that was discretized), directly to `Zr`
escalc(measure = "ZBIS",
      m1i = 13.5, m2i = 10.1,
      sd1i = 1.5, sd2i = 2.1,
      n1i = 15, n2i = 15)

```
