---
title: "Continuous predictors: assumptions and interpretation"
subtitle: "Fisher's Z-transformed correlation coefficient ($Z_r$)"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**On this Page**"
    output-file: "2.3-Zr-assumptions.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: 
  - references.bib
---

<!-- csl: ecology.csl -->

```{r}
#| eval: true
#| echo: false
#| out-width: "70%"

rm(list = ls())

if (!requireNamespace("metatools")) {
    pak::pak("ejlundgren/metatools")
}
library("metatools")

library("ggplot2")
library("data.table")
library("ggpattern")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")
```

As you have seen so far in this guide, effect sizes are not trivial things. Together, point estimates and sampling variances are small statistical models with their own numerical assumptions, properties, and quirks. Understanding these assumptions is essential if you want to conduct a proper and robust meta-analysis, and if you want to correctly interpret these effect sizes and the outputs of meta-analytic models.

In this section, we discuss effect sizes for describing the relationship strength between two continuous variables. These are known as measures of association. While the Pearson product-moment correlation coefficient ($r$) is the most intuitive and widely reported metric in primary studies, its mathematical properties make it problematic for direct meta-analytic pooling. Instead, we will dive into the elegant statistical world of **Fisher's Z-transformed correlation coefficient** ($Z_r$), our variance-stabilizing alternative.

---

### Why can't we just pool raw correlations?

When primary studies report their results, they often present Pearson's correlation coefficient ($r$). It is an appealing effect size: it carries no units, so a correlation between body mass and clutch size can be compared directly against one between soil nitrogen and plant diversity (if that makes sense to be compared, which is another question). It runs from $-1$ (a perfect negative linear association) through $0$ (no *linear* association) to $+1$ (a perfect positive linear association).

Convenient as it is, the raw correlation coefficient has three mathematical properties that sit awkwardly with the assumptions of standard meta-analytic models. All three trace back to a single issue: $r$ is bounded by $-1$ and $+1$, and statistical estimates behave badly near hard limits.

1. **The estimate gets squeezed against the ceiling.** When the true correlation in the population is near zero, repeated studies scatter symmetrically around it. When the true correlation is strong, say $+0.8$, there is simply less room to overshoot than to undershoot, because no estimate can exceed $+1$. The spread of plausible estimates becomes lopsided, with a long tail on the low side. The practical cost is that confidence intervals calculated the usual way assume a symmetric spread, so they end up miscalibrated and can even extend beyond $1$, which does not make sense. Separately, and through a different route, $r$ is a slightly *shrunken* estimate of the true correlation. It is pulled toward zero, whatever its sign. That shrinkage is small (for a true correlation of $0.5$ measured on $20$ individuals, roughly $0.01$), but it is the reason unbiased alternatives to $r$ exist [@Olkin1958-op].

2. **Precision is tangled up with the result.** Meta-analysis weights each study by its precision, and this only works if precision is governed by something independent of the answer, such as sample size. For correlations it is not. The sampling variance of $r$ is

$$
v_i = \frac{(1 - r_i^2)^2}{n_i - 1}
$${#eq-r-variance}

which contains the correlation itself. The larger the correlation a study happens to report, the smaller its variance and the heavier its weight. A study that overestimated the correlation by chance is therefore rewarded with extra influence, and the pooled estimate is dragged away from zero. Notice that this pushes in the *opposite* direction to the shrinkage in point 1, above. The same entanglement distorts funnel plots and inflates the false-positive rate of Egger's test for funnel asymmetry. The scatter narrows toward large correlations for reasons that have nothing to do with genuine precision.

3. **Small samples make matters worse.** Meta-analytic models assume each study's estimate is drawn from a bell-shaped distribution around the true value. For correlations, that approximation holds up well with large samples and modest effects, and it breaks down when the sample is small *and* the correlation is strong—the combination most familiar to anyone working with a dozen populations or a handful of species.

To solve these issues, Fisher proposed a variance-stabilizing, normalizing transformation [@Fisher1921-zr; @borenstein2021].

---

### The statistical mechanics of $Z_r$

#### 1. Fisher's $Z$ transformation (the point estimate)

Fisher's transformation stretches the bounded $[-1, +1]$ interval across the entire number line, pushing the ceilings out to infinity so that estimates are no longer squeezed against them. Formally, it is the inverse hyperbolic tangent ($\text{atanh}$):

$$
y_i = Z_{r_i} = \text{atanh}(r_i) = \frac{1}{2} \ln\left(\frac{1 + r_i}{1 - r_i}\right)
$${#eq-zr-point}

Correlations near zero barely move. Correlations near the boundaries are stretched dramatically. The result is a distribution that is close to symmetric and close to normal for samples of roughly $10$ or more.

#### 2. The sampling variance

The real beauty of Fisher's transformation lies in its sampling variance. After transformation, the sampling variance depends on sample size and nothing else:

$$v_i = \text{Var}(Z_{r_i}) = \frac{1}{n_i - 3}$${#eq-zr-variance}

The effect size has vanished from the variance entirely. Studies are now weighted purely by how much data they collected, which is exactly the assumption meta-analytic models were built on. Symmetry and approximate normality come along for free.

::: callout-note
##### The variance formula assumes bivariate normality

Fisher's variance is derived assuming the two variables follow a bivariate normal distribution. When the underlying data are strongly non-normal (heavy tailed) abundance counts, or ratios with a few extreme values, the true sampling variance can be noticeably larger than $1/(n_i - 3)$, and standard errors will be too optimistic [@Hawkins1989-zr; @Kowalski1972-nn]. This is a further reason to prefer robust variance estimation (see below) when your evidence base is built on skewed ecological data.
:::

#### 3. Back-transformation to $r$

We conduct all statistical modeling, random-effects estimation, and moderator tests in the $Z_r$ metric, but $Z_r$ values are difficult to interpret conceptually. An effect size of $Z_r = 1.1$ means very little to our brain until we convert it back to $r = 0.8$. We therefore back-transform final pooled estimates and interval boundaries to the intuitive correlation scale using the hyperbolic tangent ($\tanh$):

$$r = \tanh(Z_r) = \frac{e^{2 Z_r} - 1}{e^{2 Z_r} + 1}$${#eq-zr-back}

::: callout-warning
##### Back-transformation is not a neutral operation

Because the conversion from $Z_r$ to $r$ is curved rather than straight, back-transforming a pooled estimate does not return the average of the underlying correlations. It returns something closer to their median, and the gap between the two widens as between-study heterogeneity grows. In ecology and evolution, where heterogeneity is usually substantial, a back-transformed mean will typically exaggerate the magnitude of the true average correlation.

Three practical consequences:

* **Interval bounds back-transform safely.** Because the transformation preserves rank order, converting the two ends of a confidence or prediction interval gives valid bounds on the $r$ scale. So, convert the endpoints, never the standard error.
* **Heterogeneity does not back-transform at all.** Report $\tau^2$, $I^2$, and prediction intervals on the $Z_r$ scale, or convert the prediction interval endpoints only. There is no meaningful "$\tau^2$ in correlation units."
* **Word your conclusion carefully.** "The pooled effect corresponds to a correlation of 0.42" is defensible; "the average correlation across studies was 0.42" is not, once heterogeneity is appreciable.

Some authors (notably the Hunter–Schmidt tradition; @Hunter2014-hs; @Field2001-cf) respond to this by pooling raw correlations with sample-size weights instead. We use Fisher's $Z_r$ throughout this guide because it solves three problems at once, but the choice is genuinely contested and worth acknowledging in your methods section.
:::

---

### Code and formula pairing: implementation in R

Let's look at how to implement these calculations in R using `metafor::escalc` and basic math.

::: callout-note
##### Fisher's $Z$ in `metafor`

The `metafor` package [@Viechtbauer2010-uq] implements Fisher's $Z$ transformation via the measure code `"ZCOR"`. This is the measure you want for the workflow in this chapter.

You may also encounter `"UCOR"`, which applies the Olkin-Pratt correction [@Olkin1958-op] for the small-sample shrinkage described above. Be aware that `"UCOR"` returns a corrected *raw* correlation, not a transformed one. So it fixes the shrinkage while leaving the variance-effect entanglement fully intact. It is not an upgraded version of `"ZCOR"`, and the two should not be mixed within a single analysis.
:::

```{r}
#| eval: true
#| echo: true

# Simple demonstration of Zr transformation and back-transformation in R

# Let's say we have extracted a correlation of r = 0.6 from a study of n = 50
r_val <- 0.6
n_val <- 50

# 1. Manual Fisher's Z transformation
z_manual <- 0.5 * log((1 + r_val) / (1 - r_val))
var_manual <- 1 / (n_val - 3)

# 2. Using escalc from metafor
dat_example <- data.frame(ri = r_val, ni = n_val)
dat_zr <- escalc(measure = "ZCOR", ri = ri, ni = ni, data = dat_example)

print(paste("Manual Z:", round(z_manual, 4)))
print(paste("escalc Z (yi):", round(dat_zr$yi[1], 4)))
print(paste("escalc Z (vi):", round(dat_zr$vi[1], 4)))

# 3. Back-transformation using tanh
r_back <- tanh(z_manual)
print(paste("Back-transformed r:", round(r_back, 4)))
```

<details>
<summary>Click here to see a visual demonstration of the transformation stretching effect</summary>

Let's plot how $r$ maps to $Z_r$. Notice how the relationship is nearly a straight line around zero but curves sharply as it approaches the boundaries of $-1$ and $+1$:

```{r}
#| eval: true
#| echo: true
r_seq <- seq(-0.99, 0.99, by = 0.01)
zr_seq <- atanh(r_seq)

plot_data <- data.frame(r = r_seq, zr = zr_seq)

ggplot(plot_data
  , aes(x = r
      ,y = zr)
  ) +
  geom_line(color = "blue", linewidth = 1) +
  theme_minimal() + 
  labs(title = "Fisher's r-to-Z transformation"
    , x = "Raw correlation (r)"
    , y = "Fisher's Z (Zr)") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray")
```
</details>

---

### Not all correlations are comparable

Before converting anything, check that the correlations you have extracted are estimating the same quantity. Three common mismatches will quietly corrupt an otherwise careful meta-analysis.

**Partial correlations are not zero-order correlations.** A study reporting the association between temperature and growth rate *after controlling for* body size is reporting a different parameter from a study reporting the raw association. Pooling the two together answers no clear question. If you must include partial correlations, flag them, model them as a moderator, and adjust the sample size to account for the covariates that were controlled for (the variance is closer to $1/(n_i - k_i - 3)$, where $k_i$ is the number of control variables; @Fisher1924-pc; @Aloe2013-pt).

**Range restriction shrinks correlations** [@Thorndike1949-rr; @Sackett2000-rr]. A study sampling across a narrow elevational band will report a weaker correlation than one spanning the full gradient, even if the underlying biology is identical. Where you can, record range or standard deviation of the predictor and use it as a moderator.

**Measurement error shrinks correlations too** [@Spearman1904-me; @Hunter2014-hs]. Noise proxies attenuate associations toward zero. If reliability estimates are available for some studies and not others, this becomes a source of heterogeneity you cannot see.

---

### Handling ranked correlations: Spearman's $\rho$ and Kendall's $\tau$

When voyaging through the literature, you will inevitably encounter studies that do not report Pearson's $r$. Instead, because their data were non-normal or ranked, they reported **Spearman's rank correlation** ($\rho$ or $r_s$) or **Kendall's tau** ($\tau$).

Because these metrics are based on ranks rather than raw values, their numeric scales are slightly compressed compared to Pearson's $r$. To include these studies in your meta-analysis, you can apply the following transformations to approximate their Pearson equivalents:

#### Spearman's $\rho$ to Pearson's $r$:

$$r \approx 2 \sin\left(\frac{\pi}{6} \rho\right)$${#eq-spearman}

#### Kendall's $\tau$ to Pearson's $r$:

$$r \approx \sin\left(\frac{\pi}{2} \tau\right)$${#eq-kendall}

Both formulas assume the underlying data are bivariate normal [@Rupinski1996-rd; @Kendall1948-rk], which is very often the assumption the original authors doubted when they chose a rank method in the first place. Treat the conversions as approximations.

A useful refinement: once converted and Fisher-transformed, an effect size originating from Spearman's $\rho$ is slightly noisier than one originating from Pearson's $r$. Because ranks discard some metric information, converted rank correlations have slightly lower statistical efficiency than Pearson correlations, inflating the sampling variance of $Z_r$ by about 6% under bivariate normality [@Fieller1957-rk; @Myers2004-rk]: 

$$v_i \approx \frac{1.06}{n_i - 3}$${#eq-rank-inflation}

::: callout-important
##### Sensitivity analyses as best practice

Never treat these conversions as magic solutions. Approximations introduce uncertainty. Always flag converted effect sizes in your dataset (*e.g.,* `is_converted = TRUE`) and **run sensitivity analyses**. Specifically, run your meta-analytic models both *with* and *without* the converted Spearman and Kendall correlations. If your overall effect size or its uncertainty shifts substantially, you must report this and be cautious in your interpretations.
:::

---

### The "unit of replication" obsession

A central theme of our guide is the **unit of replication**. When calculating and meta-analyzing measures of association, ensuring that the unit of replication aligns across studies is paramount. Mismatches in the level of replication will alter your sample size, your sampling variance, and consequently how much influence each study has on your pooled result.

Consider a study examining the correlation between soil nitrogen and plant diversity:

* **Study A** measures the correlation across **50 individual quadrats** ($n = 50$).
* **Study B** measures the correlation across **5 distinct geographic sites**, where each site represents the pooled average of 10 quadrats ($n = 5$).

Even if both studies report an identical correlation of $r = 0.5$, Study A carries roughly 47 units of weight against Study B's 2. A difference of more than twentyfold (!), driven entirely by a decision about what counts as a replicate. If site is the true independent unit of replication, treating Study A's quadrats as independent replicates is pseudoreplication [@Hurlbert1984-pr].

::: callout-note
##### Weights are not simply $n - 3$

The figures above assume a common-effects model, where a study's weight is one over its sampling variance. In the random-effects and multilevel models you will actually fit, weight is one over the sampling variance *plus* the between-study variance $\tau^2$. That extra term acts as a floor on everyone's uncertainty and compresses the differences between studies considerably: with a moderate $\tau^2$ of $0.1$, the twentyfold gap above shrinks to roughly fivefold. Large studies still dominate, but far less absolutely than the raw sample sizes suggest.
:::

Before running your models:

1. Identify the true, independent unit of replication for each study.
2. Standardize or aggregate correlations where necessary to avoid inflating sample sizes.
3. When in doubt, perform sensitivity analysis by adjusting sample sizes to reflect conservative biological units of replication.

---

### Non-independence, matrices, and tagging

Correlation coefficients frequently violate the assumption of independence. Two main traps exist in continuous predictor meta-analyses:

#### 1. Pairwise correlation matrices (the melting trap)

Primary studies often present a correlation matrix comparing multiple traits or environmental factors. Meta-analysts sometimes "melt" these matrices and include all pairwise correlations as separate effect sizes.

::: callout-warning
##### Melted matrices cause non-independence that grows quadratically

If a study measures 5 traits on a single set of individuals, it generates $\frac{5 \times (5-1)}{2} = 10$ pairwise correlations. Those 10 correlations share the exact same subjects. Treating them as independent data points inflates your Type I error rate, because the model behaves as though you had ten times more information than you actually collected.
:::

To handle this non-independence:

* Use **multilevel meta-analysis** with a random effect for study or paper identity to absorb shared variance [@Noble2017-wv].
* Where the same individuals contribute to several correlations, construct an explicit variance-covariance matrix with `metafor::rcalc` rather than relying on random effects alone.
* Combine either approach with **robust variance estimation** [@Hedges2010-rve; @Pustejovsky2018-cs] (`metafor::robust`, or the `clubSandwich` package) so that your standard errors remain trustworthy even if the assumed covariance structure is imperfect.

#### 2. Spatial and temporal flagging

Continuous predictors in ecology and evolution often vary across spatial and temporal grains. Always record and flag metadata for: 

* **Spatial grain:** *e.g.,* local plot size, regional $\text{km}^2$.
* **Temporal grain:** *e.g.,* seasonal correlation, decadal trend.
* **Flag groupings:** flag studies sharing common study sites, common control subjects, or common experimental treatments.

---

### Interpretation and benchmarks: what does $r$ mean?

When interpreting your pooled effect size, you may encounter the traditional benchmarks proposed by Jacob Cohen [-@Cohen1988-pw]:

| Effect size class | Correlation coefficient ($r$) | Variance explained ($r^2$) | Interpretation |
| :--- | :--- | :--- | :--- |
| **Small** | $0.1$ | $1\%$ | Subtle effect, barely visible to the naked eye. |
| **Medium** | $0.3$ | $9\%$ | Visible to a careful observer; typical of many social and ecological processes. |
| **Large** | $0.5$ | $25\%$ | Prominent relationship; easily detectable. |

These labels were offered by Cohen as a fallback for researchers with no domain knowledge to draw on, and he said so explicitly. The interpretation of effect sizes is highly context-dependent:

* In **evolutionary ecology**, natural selection operates on tiny margins. A correlation of $r = 0.05$---explaining a quarter of one percent of the variance---can be strongly adaptive and shape survival over generations [@Kingsolver2001-am; @Nakagawa2012-ee].
* In **biomedical research**, regular aspirin consumption has an effect on heart attack risk equivalent to $r = 0.034$. This explains about one tenth of one percent of the variance, yet translates to thousands of lives saved annually [@Rosenthal1990-ap].

Never apply Cohen's benchmarks blindly. Always interpret your pooled correlations in the context of the biology, the decision costs, and the practical implications of the phenomenon.

---

### Recommended workflow for continuous predictors

To ensure your meta-analysis is rigorous, stick to the following workflow:

1. **Extract raw data:** Record the correlation, its sample size, the type of correlation reported, whether it is zero-order or partial, and the unit of replication.
2. **Screen for comparability:** Set aside or flag partial correlations, and note any study with obvious range restriction or unreliable measurement.
3. **Standardize conversions:** Convert any Spearman's $\rho$ or Kendall's $\tau$ to Pearson's $r$ equivalents. Flag these.
4. **Calculate Fisher's $Z_r$:** Transform raw and converted correlations using `metafor::escalc(measure = "ZCOR")`.
5. **Address non-independence:** Check for melted matrices and shared subjects. Nest non-independent effect sizes within study using multilevel random effects.
6. **Run models:** Fit multilevel random-effects models in the $Z_r$ metric, paired with robust variance estimation.
7. **Test sensitivity:** Re-run models excluding converted metrics, small-sample outliers, or pseudoreplicated studies.
8. **Back-transform for reporting:** Convert pooled estimates and interval endpoints back to $r$ using `tanh`. Keep heterogeneity statistics on the $Z_r$ scale, and word your conclusions to reflect what the back-transformed value actually represents.
