---
title: "Continuous predictors"
subtitle: "The heart of observational data synthesis"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**Page Contents**"
    output-file: "2.3-Zr-assumptions.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
csl: ecology.csl
---

```{r}
#| eval: true
#| echo: false
#| out-width: "70%"

rm(list = ls())

if(!requireNamespace("metatools")) {
  pak::pak("ejlundgren/metatools")
}
library("metatools")

library("ggplot2")
library("data.table")
library("ggpattern")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

```

Zr is the dominant effect size for describing correlations. In particular, cases with a continuous predictor variable and one or two continuous response variables. Zr takes the correlation coefficient ($r$) and sample size as its inputs. The point estimate itself is the arc-tangent of the correlation coefficient (which must be bounded between -1 and 1).

$$\begin{aligned}
&r_{i} \leftarrow \min(1, \max(-1, r_i))\\
&Z_{r_i} = atanh(r_{i})\\
\end{aligned}$${#eq-1}

The sampling variance of Zr depends entirely on sample size and is defined as:

$$\begin{aligned}
&Var(Z_r) = \frac{1}{n-3} \\
\end{aligned}$${#eq-2}

As you can see, sampling variance is only calculable if $n>3$.

In `escalc`, this looks like:

```{r}
escalc(ri = 0.7, ni = 15,
       measure = "ZCOR")
```

Zr has among the least biased of point estimate and variance formulas (REF).

*What to do about 0/1 or binary outcomes with a continuous predictor?*

:::callout-question
**Zr needs Pearson's correlation coefficient not Spearmans** NEED TO MENTION IN REPORTING OF CORRELATIONS OF DIFFERENT TYPES.

Also should remind about different correlation types? Pearson's (2 continuous variables), Spearman's (1? non-parametric variable), Kendall (???), point-biserial (1 variable truly naturally discrete), biserial (1 variable artificially discretized), techronic (2 variables artificially discretized)

See @Jacobs2017-cv
 
See correlation confusion.png in the `notes/` directory
:::

## Assumptions {#sec-assumptions}

Presumably need to be truly continuous, e.g., not ordinal? Can you use directly reported correlation coefficients that aren't Pearson's??

## Interpretation {#sec-interpretation}

Zr effect sizes describe the strength and direction of a relationship between two continuous variables.

## Other correlation types {#sec-other-correlation-types}

### Point-biserial correlations & biserial correlations {#sec-point-biserial}

Point-biserial and biserial correlations are oddballs. They are very similar to SMD, in that they are an effect size that summarizes the relationship between a discrete predictor and a continuous outcome variable. However, these effect sizes are a measure of association/correlation and range from -1 to 1, instead of describing the difference in magnitude and direction. 

Point-biserial correlations are for truly distinct predictor variables (e.g., male vs female) while biserial correlations are for continuous variables that have been turned into discrete variables (e.g., large versus small or presence versus absence).

**Shouldn't use male vs female as example of truly discrete variables**

According to @Jacobs2017-cv, you should not convert point-biserial correlations to correlation coefficients. However, you can convert biserial correlations. `escalc` provides a shorthand to directly convert these statistics to Zr

```{r}

# For the biserial correlation coefficient (for a continuous predictor that was discretized), directly to `Zr`
escalc(measure = "ZBIS",
      m1i = 13.5, m2i = 10.1,
      sd1i = 1.5, sd2i = 2.1,
      n1i = 15, n2i = 15)

```

In our mind, the biggest advantage of calculating biserial correlations is that they require fewer assumptions to *convert* to Zr. For example, if you are doing a meta-analysis where the majority of your data are correlations except for one or two studies that use discrete predictors, it may be better to calculate biserial correlations for these studies instead of calculating SMD and then [converting](4-converting-effect-sizes.qmd) SMD to Zr. 

