---
title: "Discrete outcomes with discrete predictor"
subtitle: "lnOR and lnRR"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**CONTENTS**"
    output-file: "3.3-lnOR-challenges.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
---

```{r}
#| eval: true
#| echo: false
#| out-width: "70%"

rm(list = ls())
library("metaDigitise")
library("ggplot2")
library("data.table")
library("ggpattern")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

# Source our custom functions:
source("scripts/functions/eff_size.R")
source("scripts/functions/convert_effect_sizes.R")

```

Discrete outcomes are a common design in ecology and evolution, though perhaps not as common as in medical trials. These types of data are meta-analyzed with lnOR or lnRR effect sizes and require either discrete numbers per group or proportions and total group size. We recommend structuring lnOR/lnRR extraction databases *wide*, with four columns, one for each of these groups.

For example, imagine you are meta-analyzing the influence of introduced predators on the breeding success of birds. You are extracting data from articles that report the proportion of monitored nests that successfully fledged on islands with and without predators. These values correspond to a contigency table that looks like this:

|                  | Fledged | Failed |
|------------------|---------|------|-
| Predator present | $a$     | $b$  |
| Predator absent  | $c$     | $d$  |

Your empty database structure for these values might look like this (ignoring all the other important columns, like latitude, longitude, article id, etc). The`fledged_predators_present`, `failed_predators_present`, `fledged_predators_absent`, and `failed_predators_absent` correspond to `a`, `b`, `c`, `d` cells of your contingency table and to the arguments required by `metafor::escalc`.


```{r}
#| eval: true
#| echo: false
# library("kableExtra")
dat <- data.table(fledged_predators_absent = rep("", 3),
                  fledged_predators_present = rep("", 3),
                  failed_predators_absent = rep("", 3),
                  failed_predators_present = rep("", 3))
knitr::kable(dat)

```
**Add values so this doesn't look so dumb...**

Here is a cheatsheet for extracting lnOR/lnRR type data: [XXXXXXX]


# The easy cases
Let's start with some ideal, dreamy lnOR type data.

## Proportion or % data with total population

```{r}
#| echo: false
#| eval: true
#| out-width: "70%"

# Simple:
dt <- data.table(mean = c(15, 31), 
                 group = c("Control", "Treatment"))
# dt <- 
ggplot(data = dt,
       aes(x = group, y = mean, fill = group))+
  geom_col()+
  scale_fill_manual(NULL,
                    values = c("Control" = "dodgerblue2",
                               "Treatment" = "indianred"),
                    labels = c("Control" = "Control (n = 100 individuals)",
                               "Treatment" = "Treatment (n = 200 individuals)"))+  
  xlab(NULL)+
  ylab("Mortality rate (% of population)")+
  theme_bw()+
  theme(panel.grid = element_blank(),
        legend.position = "none")
```

These are also straightforward. Extract the percent and reconstruct the number of individuals in each outcome category based on total n.

## No control group

Some experiments only report the experimental treatment, assuming that the control has a probability of 50%. In these cases, you'll have to assume the same to calculate lnOR, e.g., by populating one of the columns with 50 / 50 values (generally with the same total N as the first row). For example, if a study only reports:

|           | Outcome A | Outcome B |
|-----------|-----------|-----------|
| Treatment | 35        | 15        |

Take the row sum ($35 + 15 = 50$) and populate the missing values ($0.5 * 50$) for the control:

|           | Outcome A | Outcome B |
|-----------|-----------|-----------|
| Treatment | 35        | 15        |
| Control   | 25        | 25        |

**ANNOYED BY HOW THE TABLES RENDER FULL WIDTH**

Observations like this should be flagged for sensitivity analyses.

Now let's look at some other ways this data is displayed.

## Aggregating multiple measurements

Some studies may report a response like survivorship over multiple years before and after some intervention (e.g., eradication of a predator, change in management policy, etc). Since the years before and after this treatment aren't pairable, the best approach is to summarize across years per treatment.

```{r}
#| eval: true
#| echo: false

dat <- data.table(year = 1990:1999,
                  treatment = c(rep("Pre", 4), 
                                rep("Post", 6)))
# dat

dat[, total_individuals := c(150, 131, 160, 171, 153, 158, 111, 95, 98, 101)]
dat[, number_deceased := c(5, 8, 4, 7, 11, 13, 12, 18, 15, 17)]
dat[, mortality_rate := round(number_deceased / total_individuals, 2)]
dat$treatment <- factor(dat$treatment, levels = c("Pre", "Post"))

ggplot(data = dat, aes(x = year, y = mortality_rate, fill = treatment))+
  geom_col()+
  scale_fill_manual(NULL,
                    values = c("Pre" = "dodgerblue2",
                               "Post" = "indianred"),
                    labels = c("Pre" = "Pre-treatment",
                               "Post" = "Post-treatment"))+ 
  xlab(NULL)+
  ylab("Mortality rate")+
  scale_x_continuous(breaks = 1990:1999)+
  theme_bw()+
  theme(panel.grid = element_blank(),
        legend.position = "bottom")


```

To extract data like this, you would need to determine the total population size for each year (which is hopefully reported) and multiply this by the proportion of that year to derive the number of individuals that lived and died per year to produce a table like this:

```{r}
#| eval: true
#| echo: false

knitr::kable(dat)

```

Then, take the sum of `total_individuals` and `number_deceased` and calculate the number living:

```{r}
#| eval: true
#| echo: true
#| output: asis

dat.sum <- dat %>% group_by(treatment) %>%
  summarize(total_individuals = sum(total_individuals),
            number_deceased = sum(number_deceased)) %>%
  mutate(number_living = total_individuals - number_deceased)

knitr::kable(dat.sum)

```

The columns `number_deceased` and `number_living` provide you the contingency table necessary to calculate lnOR.

However, what do you do if there is no sample size provided (e.g., total number of individuals)? In this case, your best bet is to calculate the mean ± SD of each year per treatment group and use this to calculate SMD. Since these values are proportions, be sure to arcsin transform these values (see @sec). The resulting effect size and sampling variance can then be converted to lnOR to be comparable to the rest of your dataset.


## Survivorship curves

*I don't know how to directly simulate the underlying data for these...*

```{r}
#| eval: true
#| echo: false

# >>> Survivorship --------------------------------------------------------

x <- seq(1:15)
a <- rweibull(n = 15, shape = 2, scale = 4)
# plot(x, rev(sort(a)))

c <- rweibull(n = 15, shape = 2, scale = 4)
# plot(x, rev(sort(b)))
a <- rev(sort(a))
c <- rev(sort(c))

a <- a - min(a)
a <- a/max(a)
# a
c <- c - min(c)
c <- c/max(c)
# b

ggplot()+
  geom_path(aes(x = x, y = a),
            color = "indianred",
            linewidth = 2)+
  geom_path(aes(x = x, y = c),
            color = "dodgerblue",
            linewidth = 2)+
  geom_point(aes(x = x, y = a),
            fill = "indianred",
            shape = 21,
            size = 3)+
  geom_point(aes(x = x, y = c),
            fill = "dodgerblue",
            shape = 21,
            size = 3)+
  scale_x_continuous(breaks = seq(1, 15, by = 2))+
  ylab("Survivorship (proportion alive)")+
  xlab("Time")+
  theme_bw()+
  theme(panel.grid = element_blank(),
        legend.position = "none")
```

In this case, the y axis is the proportion ($p1$ and $p2$) for each group (group 1 and 2). The meta-analyst needs to extract the total number of individuals in groups 1 (blue) and 2 (red) at time point 1, which can then be multipled by the total number of individuals at the beginning of the experiment. This will yield $a$ and $c$ for the two groups. $b$ and $d$ can then be found by subtracting from the total $n$:

|                | time | $p$       | N        | Alive   | Dead   |
|----------------|------|-----------|----------|---------|--------|
| Group 1 (blue) | 3    | $p1$=0.71 | $n1$=150 | $a$=107 | $b$=43 |
| Group 2 (red)  | 3    | $p2$=0.46 | $n2$=150 | $c$=69  | $d$=81 |

As one can see, this type of data also poses a classic time-series challenge, which can face all effect sizes. We recommend extracting the data for each time point and then using random effects to control for temporal autocorrelation in your meta-analytic model (see sec-XXXX). An alternative is to extract the time point with the greatest difference between control and treatment and analyze that.

## Mean±error outcome for two discrete groups

Sometimes discrete outcomes are summarized into mean±error across trials/sites/islands/etc. For example, a paper that monitored mortality rates across a discrete number nests on separate islands, calculated the mortality rate, and then summarized by mean±error across islands.

This data format lends itself naturally to SMD/lnRoM/Zmu type effect sizes @sec. However, the sample size for the mean±error in these summaries are the number of trials/sites/islands/etc instead of the number of individuals (or nests). If the unit of replication of the rest of your dataset is the number of individuals (or nests) then this effect size would have vastly less weight in your models (and probably shouldn't be included).
```{r}
#| echo: false
#| eval: true
#| out-width: "70%"

# Simple:
dt <- data.table(mean = c(0.15, 0.31), 
                 group = c("Control", "Treatment"),
                 se = c(0.01, 0.015))
# dt <- 
ggplot(data = dt,
       aes(x = group, y = mean, fill = group, ymin = mean-se,
           ymax = mean+se))+
  geom_col()+
  geom_errorbar(width = 0.5)+
  scale_fill_manual(NULL,
                    values = c("Control" = "dodgerblue2",
                               "Treatment" = "indianred"))+  
  xlab(NULL)+
  ylab("Mean mortality rate across sites (±standard error)")+
  theme_bw()+
  theme(panel.grid = element_blank(),
        legend.position = "none")

dt[, n_sites := ifelse(group == "Control", 7, 9)]

```

This data required a total number of individuals in each fate category (e.g., alive vs dead) and is thus ultimately best suited for lnOR type effect sizes—yet it's presented in a way that is most readily extractable for SMD/lnRoM/Zmu effect sizes. Moreover, as you can see, this figure presents standard error, not standard deviation. Converting standard error to standard deviation requires the sample size used by the study authors to calculate standard error in the first place, which is the number of sites.

In this case, we'd recommend reconstructing the fate of individuals within each site, using properties of the binomial distribution. This requires the following information:

**Is this the best option???**

1) the number of sites used in calculating the standard error.

2) the number of individuals at each site.

The number of sites is, in binomial speak, the number of trials (usually denoted $m$ but here we'll denote it as $t$ to avoid confusion with mean). The number of individuals is the number of observations ($n$).

Let's imagine that we found that the number of sites was 7 for the control treatment and 9 for the treatment. We start by using these to convert the extracted standard error ($se$) to standard deviation ($sd$) in our data.table we extracted from the figure:

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

dt[, sd := se * sqrt(n_sites)]

knitr::kable(dt)

```

**I'm not sure this is right...**

We now use the mean ($m$) and standard deviation to calculate $p$, the parameter that describes the shape of the binomial distribution. Note that the the standard deviation must be less than the square root of the mean.

$$
p=1-\frac{sd^2}{m}
$$
We also gleaned from the text that the number of individuals was 30 per site Things get more tricky, coding wise, if the number of individuals is different per site. We use this, along with `p` to simulate the underlying number of individuals that are alive vs dead across all sites per control and treatment.

`rbinom`, which we'll use to draw binomial data, takes `p` as well as `n`, which is the number of sites, and `size`, which is the number of individuals per site
**Is that correct??**

Since we're interested in the total number that lived, we take the sum of each vector produced by `rbinom`. This vector will be of length equal to the number of sites, with an integer that is the number that lived of 30 individuals per site.

**p becomes equal in both treatments...Think there must be a relationship between mean and sd that I don't understand. Or my equation is wrong...**
```{r}
dt[, p := (1 - (sd^2)/mean)]

for(i in 1:nrow(dt)){
  dt$n_alive[i] <- rbinom(n = dt$n_sites[i],
                     size = 30,
                     p = dt$p[i]) |> sum()
}

# Now calculate total number individuals:
dt[, total_individuals := 30 * n_sites]
dt[, n_dead := total_individuals - n_alive]

knitr::kable(dt)

```

Now you have the ingredients to calculate lnOR or lnRR. The sample size went from 7 and 9 to the actual number of individuals, making the unit of replication comparable with the other effect sizes in your dataset.

Note, that if there were variable numbers of individuals per site, then you'd want to create a long dataframe (e.g., by merging your extracted summaries with a table of number of individuals per site). Then loop through this longer data frame, using the number of individuals per site as your `size` in the code above. 

Let us know if you'd like us to write sample code to do something like that :)

## LC50?

Many laboratory studies, for example on the effects of a pollution agent on mortality, report LC50, which is the dosage at which 50% of the exposed individuals experience an outcome (e.g., die).

These figures look like this:

```{r}
#' @Kyle?
```

LC50 is already its own formula. We do not know how to convert this to lnOR or lnRR. If you're faced with lots of this data, you could consider converting survivorship curves, or other data sources, to LC50—and deriving a sampling variance formula or using SAFE to simulate sampling variance.

## Model estimates

According to Gemini, you can multiply an *unstandardized* logistic regression coefficient by 1.81 to convert it to lnOR. But not sure if that's true—and not sure how to calculate sampling variance.
