---
title: "Less common effect sizes"
subtitle: "Beyond the mean"
date: today
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**CONTENTS**"
    output-file: "5.0-less-common-effect-sizes.qmd"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
csl: ecology.csl
---

The effect sizes that we've described in depth in this tutorial have concerned the direction and magnitude of *mean* relationships between two variables.

However, there are also effect sizes to describe the absolute magnitude of differences between groups (**lnM**), to describe differences in variability (or heteroscedasticity) between two groups (**lnVR** and **lnCVR**) and effect sizes to summarize single-group means and proportions. Extracting data for these effect sizes should be fundamentally similar as the other effect size types that this tutorial was dedicated to. We here briefly discuss these effect sizes, their assumptions, and how to calculate them in `R`.

# Differences between groups {#sec-differences-between-groups}

## Differences in magnitude {#sec-lnM}
**lnM** is an effect size that describes the absolute difference between two groups, which is often relevant in evolutionary research where the direction of difference between two populations or species is not of interest relative to absolute magnitude of differences [@Nakagawa2026-wn]. This effect size replaces earlier efforts which took the absolute value of SMD (or 'folded-normal SMD'). 

To calculate lnM, you need the sample sizes ($n_1$, $n_2$), means ($m_1$, $m_2$), and standard deviations ($sd_1$, $sd_2$) for two discrete groups, similar as when calculating SMD or lnRoM effect sizes. Note that there is also a paired version of lnM that requires only 1 sample size ($n$) and $rho$---the estimated correlation between the two groups. As with other [paired effect sizes](2.1-SMD-assumptions.qmd#sec-paired-designs), since $rho$ is usually unreported, you can use values of 0.5 or 0.8 (and conduct sensitivity analyses!).

### Assumptions {#sec-lnM-assumptions}
Blah blah blah. SANTI

### Interpretation {#sec-lnM-interpretation}
Complicated and weird. Need to subtract something from it. Blah blah blah. SANTI

### How to calculate {#sec-lnM-how-to}
The math of lnM can lead to uncalcaluable values when the means of the two groups are similar. For this reason, this effect size should use SAFE bootstrapping [@Nakagawa2025-do].

As such, this effect size has to be calculated in the package `orchaRd` [@Nakagawa2023-ol] or with our function `eff_size`:

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

source("scripts/functions/eff_size.R")
source("scripts/functions/convert_effect_sizes.R")

library("orchaRd")

```

In `eff_size` be sure to set `SAFE=TRUE` in the function arguments:

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

eff_size(x1 = 15, x2 = 10, 
         sd1 = 1.5, sd2 = 2,
         n1 = 15, n2 = 15,
         effect_type = "lnM",
         SAFE = TRUE)

```
The `yi` and `vi` are the plugin-calculations of lnM point estimate and sampling variance, respectively. The `yi_safe`, `vi_safe`, and `SE_safe` are SAFE calculations. To learn more about SAFE see the tutorial [here](https://ejlundgren.github.io/SAFE/) and @Nakagawa2025-do. 

For the paired version of lnM, add `paired = TRUE` to the `eff_size` function call above.

In `orchaRd`, use the internal function `.safe_lnM_indep` (for independent, unpaired cases). Note that this has to be namespaced with three colons (`:::`):

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

orchaRd:::.safe_lnM_indep(x1bar = 15, x2bar = 10, 
                         sd1 = 1.5, sd2 = 2,
                         n1 = 15, n2 = 15)

```
This function returns both the point estimate (`lnM_SAFE`), variance (`var_lnM_SAFE`) and the number of successful SAFE bootstraps.

And for dependent cases:
```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

orchaRd:::.safe_lnM_dep(x1 = 15, x2 = 10, 
                       sd1 = 1.5, sd2 = 2,
                       n = 15, r = 0.5)

```

Please see the following tutorial for more information about lnM [link](https://itchyshin.github.io/meta-analysis_of_magnitude/).

## Differences in variability {#sec-two-group-variability}
The most commonly used effect sizes concern *mean* relationships. However, *variability* is itself a meaningful dimension of reality, not merely noise. In ecological contexts, heterogeneity is considered a key driver of biodiversity and ecosystem functioning [@Stein2014-gj; @Levine2016-hb]. Likewise, in evolution, variability tells us important information about selective pressures which can either act to increase variation [@Buckley2016-sn; @O-Dea2016-jy] or reduce variation [@Crispo2007-nf]. 

There are two main effect sizes for calculating variability or heteroscedasticity between two groups (i.e., discrete predictors): **lnVR** and **lnCVR** [@Nakagawa2015-tq]. 

### lnVR {#sec-lnVR}
**lnVR** is the *log-transformed variability ratio*, defined as $log(\frac{sd_1}{sd_2})$. This effect size is **not** standardized by the mean. For that reason, this effect size can be applied to measurements derived from groups with a mean ≤ 0. However, the means need to be directly comparable. 

This can be calculated in `escalc` with the measure = "VR".

```{r}
#| eval: false
#| echo: true
#| out-width: "70%"
escalc(measure = "VR",
       sd1i = 1.5, sd2i = 2.5,
       n1i = 15, n2i = 15)
```

There is a paired version of lnVR as well, which requires an estimate of the correlation between group 1 and 2, which we can specify as 0.5 or 0.8 if it's unreported (which is typical, see [here](2.1-SMD-assumptions.qmd#sec-paired-designs)):

In `escalc` this is specified with the "VRC" measure:

```{r}
#| eval: false
#| echo: true
#| out-width: "70%"
escalc(measure = "VRC",
       sd1i = 1.5, sd2i = 2.5,
       ni = 15, ri = 0.5)
```

### lnCVR {#sec-lnCVR}
**lnCVR** is the *log-transformed coefficient of variation ratio*, defined as $log(\frac{sd_1/m_1}{sd_2/m_2})$. This effect size **is** standardized by the mean allowing you to compare groups with different means. However, this effect size cannot accept negative values in the mean, requiring positive, ratio-scale data like the lnRoM effect size.

This can be calculated in `escalc` with the measure = "CVR". This effect size requires the same statistical ingredients as SMD and lnRoM:

```{r}
#| eval: false
#| echo: true
#| out-width: "70%"
escalc(measure = "CVR",
       m1i = 15, m2i = 15,
       sd1i = 1.5, sd2i = 2.5,
       n1i = 15, n2i = 15)
```


There is a paired version of lnCVR as well. In `escalc` this is specified with the "CVRC" measure:

```{r}
#| eval: false
#| echo: true
#| out-width: "70%"
escalc(measure = "CVRC",
        m1i = 15, m2i = 15,
       sd1i = 1.5, sd2i = 2.5,
       ni = 15, ri = 0.5)
```

# Single group effect sizes {#sec-single-group}
While most meta-analyses concern the relationships between two variables, often reflecting experimental treatments, sometimes we're interested in summarizing single-group measurements and then analyzing these measurements in relation to moderators.

For example, imagine you want to understand the drivers of predator attacks on livestock. Various studies have reported predator diets or kill frequencies, allowing you to obtain the proportion of diet that contains livestock. These single-group proportions can be converted to effect sizes (so that they obtain a sampling variance estimation). You can then use metaregression to evaluate the influence of various moderators (e.g., human density, elevation, prey density) on livestock depredation rates.

:::callout-important
In some situations, even experimental data may potentially, be better to analyze as a single-group effect size if most studies used >2 experimental groups or if experimental treatments were extremely different than each other. 

For example, instead of calculating two-group effect sizes between each experimental treatment and the control (e.g., SMD, lnRoM, or lnOR, lnRR), you could calculate single group effect sizes for each group, and then use the various experimental groups as predictors in a metaregression. 

This would be especially useful if the experimental treatments were different densities of an organism (say a predator or herbivore) or different temperature treatments. 

Note that this is possible both with continuous data summarized by mean±sd as well as with binary outcome data.
:::

## Single group standardized mean {#sec-standardized-mean}
The *single group standardized mean* is the mean divided by standard deviation to create a unit-free index, similar to SMD. As such, this effect size can accept values ≤ 0.

In `escalc`:
```{r}
#| eval: false
#| echo: true
#| out-width: "70%"

escalc(measure = "SMN",
       mi = 10,
       sdi = 1.5,
       ni = 15)
```
This, like SMD, can be interpreted as the number of standard deviations the mean is from 0. Given that standard deviation influences the mean, this effect size may be less preferred with overdispersed data (as with SMD). See [here](2.1-SMD-assumptions.qmd#sec-overdispersion), where we discuss how to diagnose overdispersion in your data.

## Log-transformed mean {#sec-log-mean}
The *log-transformed mean* is a single group corollary to lnRoM, just as the *single group standardized mean* is a single group version of SMD. This effect size cannot accept negative values. [As with lnRoM](2.1-SMD-assumptions.qmd), the point estimate for log-transformed mean is not influenced by standard deviation, which is instead accounted for by the sampling variance estimate.

In `escalc`:
```{r}
#| eval: false
#| echo: true
#| out-width: "70%"

escalc(measure = "MNLN",
       mi = 10,
       sdi = 1.5,
       ni = 15)
```

As such, this effect size is likely better suited for [overdispersed](2.1-SMD-assumptions.qmd#sec-overdispersion) data. 

## Single group variability {#sec-single-group-variability}
Much as we can meta-analyze differences in variability between two groups with the lnVR and lnCVR effect sizes, we can also summarize the variability of single groups with the *log-transformed standard deviation* and *log-transformed coefficient of variation* effect sizes. 

As with [lnVR](#sec-lnVR), *log-transformed standard deviation* is not standardized by the mean and thus must be compared to studies using the same units of measurement. 

To calculate *log-transformed standard deviation* in `R`:
```{r}
#| eval: false
#| echo: true
#| out-width: "70%"

escalc(measure = "SDLN",
       sdi = 1.5,
       ni = 15)
```

The *log-transformed coefficient of variation* is standardized by the mean, just like [lnCVR](#sec-lnCVR), and can thus be used to compare data with different units of measurements. In `R`:
```{r}
#| eval: false
#| echo: true
#| out-width: "70%"

escalc(measure = "CVLN",
       sdi = 1.5,
       mi = 15,
       ni = 15)
```


## Single group proportions {#sec-proportions}
The single group measurements so far pertained to normal data (or at least quasi-normal data). However, there are also single group effect sizes concerned with the probability of binary outcomes (e.g., binomial data). These are the single group corollaries to [lnOR and lnRR](2.2-lnOR-assumptions.qmd) type effect sizes.

These effect sizes take into account the total number of individuals and the number of individuals that experienced the binary outcome (e.g., mortality). There are several options for single group proportional data like this in `escalc` of which it appears that the *log-transformed proportion*, *logit-transformed proportion (log odds)*, and *probit-transformed proportion* appear to be the most numerically stable.

These different versions have different numeric assumptions

**I'm basically taking this from Gemini, which I really don't like doing....because it's so often wrong**

- *log-transformed proportion*: This effect size appears to be best when proportions are low (e.g., p < 0.2) and when values rarely reach 1. Values of 0 require the addition of a small constant. Model estimates produced with this effect size can be exponentiated back to proportion scale for interpretation (e.g., with `exp(value)`). Specified with "PLN" in `escalc`. 

- *logit-transformed proportion*: This is the best choice when proportions are between 0.2 and 0.8. Proportions of 0 and 1 require data correction (e.g., adding or subtracting a small constant). This effect size is interpreted as log odds, the same as for [lnOR](2.2-lnOR-assumptions.qmd#sec-lnOR-interpretation) after exponentiating (`exp(value)`). Specified with "PLO" in `escalc`.

- *probit-transformed proportion*: This effect size is often used for toxicological or dose-response models, like [LC50](3.3-lnOR-challenges.qmd#sec-lc50), and is based on modeling cumulative probabilities. This effect size is less intuitive to interpret: values of 0 indicate that the proportion is 50%, while positive values indicate that the proportion is greater than 50% and negative values less than 50%. Specified with "PRZ" in `escalc`.

We recommend that in many cases, the *logit-transformed proportion* is the effect size choice with proportion data.

For all of these effect sizes, `xi` is the number of individuals experiencing the event of interest and `ni` is the total number of individuals:

```{r}
#| eval: false
#| echo: true
#| out-width: "70%"

escalc(measure = "PLO",
       ni = 15,
       xi = 10)
```

## Event counts {#sec-event-counts}
**I'm not sure I understand this versus binary outcomes...It seems to take the same exact inputs except `ti` = total person-times at risk (equivalent to `ni`)...**



