---
title: "Variability between groups"
subtitle: "Beyond the mean"
date: today
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**On this Page**"
    output-file: "5.1-variability-between-groups.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
csl: ecology.csl
---


```{r}
#| eval: true
#| echo: false

rm(list = ls())

if(!requireNamespace("metatools")) {
  pak::pak("ejlundgren/metatools")
}
library("metatools")

library("ggplot2")
library("data.table")
library("ggpattern")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

```

**PASTED TOGETHER CONTENT FROM SEPARATE PAGES. NOT EDITED**

The most commonly used effect sizes concern *mean* relationships between variables. 

However, *variability* is itself a meaningful dimension of reality, not merely noise [@Nakagawa2026-wm]. In ecological contexts, heterogeneity is considered a key driver of biodiversity and ecosystem functioning [@Stein2014-gj; @Levine2016-hb]. Likewise, in evolution, variability tells us important information about selective pressures which can either increase variation [@Buckley2016-sn; @O-Dea2016-jy] or reduce it [@Crispo2007-nf]. 

There are two main effect sizes for calculating variability between two groups (i.e., discrete predictors): **lnVR** and **lnCVR** [@Nakagawa2015-tq]. 

## lnVR

lnVR is the log-transformed ratio of the standard deviations of two groups:

$$
\mathrm{lnVR}=\ln\left(\frac{sd_1}{sd_2}\right),
$${#eq-lnvr}

where $sd_1$ is the treatment-group standard deviation and $sd_2$ is the control-group standard deviation. If group 1 is your treatment or experimental group:

- $\mathrm{lnVR}>0$: the treatment group is more variable than the control group;

- $\mathrm{lnVR}=0$: the groups have equal standard deviations;

- $\mathrm{lnVR}<0$: the treatment group is less variable.

#### Interpretation 
lnVR can be back transformed to plain english by exponentiating (i.e., $e^{lnVR}$). 

If we have an lnVR of 0.4: $e^{0.4}\approx1.5$. This indicates that the standard deviation of the treatment group is approximately 50% higher than the control group.

Combining an lnRoM or SMD meta-analysis of mean differences with an lnVR meta-analysis can provide a more complete biological description of biological change.

#### How to calculate 
To calculate lnVR, use `measure="VR"` in `escalc`:
```{r}
escalc(measure = "VR",
       sd1i = 3, n1i = 20,
       sd2i = 2, n2i = 20)
```

There is a paired version of lnVR as well, which requires an estimate of the correlation between group 1 and 2, which we can specify as 0.5 or 0.8 if it's unreported (which is typical, see [here](2.1-SMD-assumptions.qmd#sec-paired-designs)). Use this formula if groups 1 and 2 are paired, e.g., from paired plots, before-after experiments, and so on.

In `escalc` this is specified with the "VRC" measure:

```{r}
#| eval: false
#| echo: true
#| out-width: "70%"
escalc(measure = "VRC",
       sd1i = 1.5, sd2i = 2.5,
       ni = 15, ri = 0.5)
```


## lnCVR {#sec-lnCVR}

**lnCVR** is the *log-transformed coefficient of variation ratio*. Its definition formula (prior to small sample bias correction) is: 

$$
\begin{aligned}
lnCVR=\ln\left(\frac{s_1/m_1}{s_2/m_2}\right)
\end{aligned}
$${#eq-lncvr}

This effect size **is** standardized by the mean. As such, this effect size cannot accept negative mean values, requiring positive, ratio-scale data like the [lnRoM effect size](2.1-SMD-assumptions.qmd). (Note that standard devations can never be negative so **lnVR** is always calculable as long as standard deviations were reported, even if means are <0).

Importantly, lnCVR describes both mean differences between groups and differences in variability and has an exact mathematical relationship with **lnVR** and **lnRoM**:

$$
lnCVR = lnVR - lnRoM
$${#eq-lncvr-lnvr-lnrom}

:::callout-important
Because **lnCVR** is defined by both **lnVR** and **lnRoM**, we recommend analyzing both lnVR and lnRoM (or SMD) separately instead of using lnCVR for most meta-analyses. Together, they will describe both differences in variability and differences in mean relationships.
:::

**lnCVR** is intended to account for a common biological pattern: groups with larger means often also have larger variances. However, dividing by the mean makes an implicit scaling assumption. In simple terms, it works most naturally when the standard deviation increases roughly in proportion to the mean. Equivalently, variance then increases roughly with the square of the mean.

That relationship differs among data types:

- **Approximately normal data with constant variance:** the standard deviation may stay similar as the mean changes. The coefficient of variation will then change mechanically with the mean.

- **Poisson count data:** variance equals the mean, so standard deviation increases with the square root of the mean rather than directly with the mean. The coefficient of variation therefore decreases as the mean increases.

- **Negative-binomial count data:** variance can increase much faster than the mean. At large means, the standard deviation may be close to proportional to the mean, but the relationship depends on the dispersion parameter.

- **Proportions and other bounded responses:** variance is constrained by the mean and commonly shrinks near the boundaries. The mean–variance relationship can therefore be strong and nonlinear.

For these reasons, do not choose lnCVR automatically. First plot or model the relationship between group means and standard deviations across the dataset. Use lnCVR when *relative variability* is the intended estimand and its assumed mean–variance scaling is plausible. In all cases, we recommend using both lnVR and lnCVR, as well as lnRoM or SMD. If the results are different, these sensitivity analyses can help you understand **why**.

#### How to calculate
This can be calculated in `escalc` with `measure = "CVR"`. This effect size requires the same statistical ingredients as SMD and lnRoM:

```{r}
#| eval: false
#| echo: true
#| out-width: "70%"
escalc(measure = "CVR",
       m1i = 15, m2i = 15,
       sd1i = 1.5, sd2i = 2.5,
       n1i = 15, n2i = 15)
```

There is a paired version of lnCVR as well. In `escalc` this is specified with `measure="CVRC"`.

```{r}
#| eval: false
#| echo: true
#| out-width: "70%"
escalc(measure = "CVRC",
        m1i = 15, m2i = 15,
       sd1i = 1.5, sd2i = 2.5,
       ni = 15, ri = 0.5)
```

