---
title: "Differences in magnitude between two groups"
subtitle: "Beyond the mean"
date: today
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**On this Page**"
    output-file: "5.0-differences-in-magnitude.qmd"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
csl: ecology.csl
---


```{r}
#| eval: true
#| echo: false

rm(list = ls())

if(!requireNamespace("metatools")) {
  pak::pak("ejlundgren/metatools")
}
library("metatools")

library("ggplot2")
library("ggtext")
library("data.table")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

```

The effect sizes that we've described in depth in this tutorial have concerned either *mean* relationships or differences in [variability](5.1-variability-between-groups.qmd). However, what if you're interested in the **magnitude** of mean differences, regardless of their direction? These types of questions are common in evolution, where absolute differences between two populations are of interest but directionality of those differences is inconsequential. Likewise, some ecological moderators (like population density) may influence magnitude of ecological effects but not their directionality.

Previous attempts to conduct meta-analyses of magnitude have used the absolute value of SMD as an effect size (or 'folded-normal SMD'). However, taking the absolute value of SMD produces effect sizes clustered just above 0 with distributions that violate expectations of normality and can prevent accurate model estimates.

**lnM** is a recently developed effect size that is unbounded and normally distributed and describes the absolute difference between two groups [@Nakagawa2026-rb]. 

To calculate lnM, you need the same statistical ingredients as for calculating **SMD** and **lnRoM** effect sizes: sample sizes ($n_1$, $n_2$), means ($m_1$, $m_2$), and standard deviations ($sd_1$, $sd_2$) for two discrete groups.

The math for **lnM** is built upon classical one-way ANOVA building blocks. Because the math is a bit long, we've hidden it in the folded block below. Click to see.

::: {.callout-note collapse="true"}

#### click here to see the mathematics of lnM

lnM is based on the classical ANOVA building blocks. First we calculate Mean Square Between ($MS_B$) and Mean Square Within ($MS_W$). $MS_B$ estimates the difference between group means. $MS_W$ estimates the variance inside of each group. 

$$
\begin{aligned}
&MS_B = \frac{n_1 n_2}{n_1 + n_2} * (m_1 - m_2)^2\\ 
&MS_W = \frac{(n_1 - 1)sd_1^2 + (n_2 - 1)sd_2^2}{n_1 + n_2 - 2}\\
\end{aligned}
$${#eq-ms}

We then establish within-group ($s_W^2$) and between-group variance ($s_B^2$), standardized by harmonic-mean sample size ($n_0$).

$$
\begin{aligned}
&n_0 = \frac{2n_1n_2}{n_1 + n_2}\\
&s_W^2 = MS_W\\
&s_B^2 = \frac{MS_B - MS_W}{n_0} \\
\end{aligned}
$${#eq-variances}

Finally, with use these ingredients to calculate **lnM** as the ratio of between group standard deviation ($s_B$) and within-group standard deviation ($s_W$):

$$
\begin{aligned}
&lnM = \ln\left(\frac{s_B}{s_W}\right)\\
\end{aligned}
$${#eq-lnM}

:::

### How to calculate lnM {#sec-lnM-how-to}
If the means of group 1 and group 2 are similar, lnM cannot be calculated algebraically because of 0's in the denominator. For this reason, lnM should be estimated with SAFE bootstrapping [@Nakagawa2025-do], which simulates hyperparameters based on summary statistics (i.e., $m_{1,2}$, $sd_{1,2}$) and thus, if repeated enough, can produce estimates of lnM. 

As such, **lnM** is not currently supported by `metafor::escalc` and has to be calculated in  `metatools::eff_size`. 

Be sure to set `SAFE=TRUE` in the function arguments. The `SAFE_boots` argument indicates how many bootstraps to run (1e6 is default and [appears](https://ejlundgren.github.io/SAFE/) to be a good compromise between accuracy and speed). `SAFE_max_secs` is the number of seconds to run before timing out per effect size calculation. Calculating **lnM** can be time consuming because calculations must be done over and over until a sufficient number of bootstraps is attained.

Let's create a dataset with 2 row for which we want to calculate lnM. Note that the first row of this data has means that are very similar (15 and 15.5) between the two groups and will thus require more time to run:

```{r}
#| eval: true
#| echo: true


if(!requireNamespace("metatools")) {
  pak::pak("ejlundgren/metatools")
}
library("metatools")


dat <- data.table(x1 = c(15, 15), x2 = c(15.5, 20),
                  sd1 = c(2, 2), sd2 = c(2, 2),
                  n1 = c(15, 15), n2 = c(15, 15))
knitr::kable(dat)

out <- eff_size(x1 = x1, x2 = x2, 
         sd1 = sd1, sd2 = sd2,
         n1 = n1, n2 = n2,
         data = dat,
         bind = TRUE,
         effect_type = "lnM",
         verbose = FALSE,
         SAFE = TRUE,
         SAFE_boots = 1e6,
         SAFE_max_secs = 15)

knitr::kable(out)

```

`yi` in the resulting table is the algebraic point estimate. You'll note that `yi` for the first row is `NA`. This is because the difference between `x1` and `x2` is very small, leading to incalculable values. 

SAFE circumvents this by bootstrapping from the means and standard deviations until a sufficient number of bootstraps is reached (defined by `SAFE_boots`, here set to 1e6). The point estimate and sampling variance produced by SAFE bootstrapping are reported as `yi_safe` and `vi_safe`. Additional information on the performance of the SAFE boostrapping are returned as well (`number_SAFE_iterations`, `number_SAFE_bootstraps_completed`, `SAFE_elapsed_time`, `SAFE_complete`). See function documentation for more information (`?eff_size`).

:::callout-note
##### Optimizing SAFE
Cases where the means between groups are very similar may take a while to successfully run. Increase `SAFE_max_secs` and reduce `SAFE_boots` until your dataset produces finalized SAFE estimates, indicated by the `SAFE_complete` column. We recommend not reducing `SAFE_boots` less than 1e4 (1e5 and 1e6 are preferable).

To learn more about SAFE see the tutorial [here](https://ejlundgren.github.io/SAFE/) and @Nakagawa2025-do. 
:::

:::callout-note
##### Paired **lnM**
For the paired version of lnM, add `paired = TRUE` to the `eff_size` function call:

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

eff_size(x1 = 15, x2 = 10, 
         sd1 = 1.5, sd2 = 2,
         n = 15, r = 0.8,
         effect_type = "lnM",
         paired = TRUE, # Add this. FALSE is default.
         SAFE = TRUE,
         SAFE_boots = 1e6,
         SAFE_max_secs = 60)


```
:::

### Interpretation {#sec-lnM-interpretation}
**lnM** is the ratio of between group variation and within-group variation. This between group standard deviation can be thought of as mean-separation---how far apart the two groups are.

However, lnM can be difficult to interpret. An lnM of 0 corresponds to **identical** between and within group variation. This should **not** be interpreted as **equal group means** but whether the between-group dispersion is at least as large as the within-group dispersion. 

While meta-analytic models should be fit on lnM directly, we recommend mapping lnM model estimates and confidence intervals to SMD, or $d$, for interpretation. We'll refer to this as $d_{eq}$ ("d-equivalent"):

$$
\begin{aligned}
&n_0 = \frac{2 n_1 n_2}{n_1 + n_2}\\
&d_{eq} = \sqrt{2e^{2lnM} + \frac{2}{n_0}}\\
\end{aligned}
$${#eq-d-eq}

Note that when lnM = 0, this translates to $d_{eq}$ of 1.41. What exactly does that mean? This means that within-group variation is large but overall mean differences between groups are 1.41 standard deviations apart. Tricky right?


However, what would you provide for $n_0$ if you want to convert model estimates and confidence intervals to $d_{eq}$? Luckily, @Nakagawa2026-rb provided a large-sample approximation for $d_{eq}$ that does not require $n_0$. This is defined as:

$$
\begin{aligned}
&d_{eq} \approx \sqrt{2} \times e^{lnM}
\end{aligned}
$${#eq-d-eq-approx}

Use this equation to convert model estimates and CIs to $d_{eq}$.

Let's illustrate lnM in `R` for three comparisons---each representing a plausible comparison of group 1 and 2. We'll draw data from the normal distribution for 3 scenarios: equivalent "true" means, and increasing means for group 2. We'll set the group standard deviations constant at 5, and change the means of group 2 by increments of 5 while holding group 1 constant at $m_1=15$. In order to clearly illustrate how lnM works, we'll set a sample size of 500 for each group---which would be an unusually high sample size for a meta-analytic dataset. We'll plot the simulated data, lnM estimates, and $d_{eq}$. We'll calculate $d_{eq}$ using @eq-d-eq because sample size is known (since these are point estimates and not model estimates).

```{r}
#| eval: true
#| echo: true
#| output: false
#| code-fold: true
#| fig-width: 10

set.seed(7)

# 
sd <- 5
n <- 500

# Group 1 distribution (fixed):
group_1 <- rnorm(n = n, mean = 15, sd = sd)

# Group 2 distributions:
distributions <- lapply(c(15, 20, 25),
                        function(x) data.table(mean_group_2 = x,
                                               group_2 = rnorm(n = n, mean = x, sd = sd))) |>
  rbindlist()

distributions[, group_1 := rep(group_1, length(unique(distributions$mean)))]

dat_summaries <- distributions[, .(x1 = mean(group_1),
                                   x2 = mean(group_2),
                                   sd1 = sd(group_1),
                                   sd2 = sd(group_2),
                                   n1 = 40,
                                   n2 = 40),
                               by = .(mean_group_2)]
dat_summaries

dat_summaries <- eff_size(x1 = x1, x2 = x2, 
                           sd1 = sd1, sd2 = sd2,
                           n1 = n1, n2 = n2,
                           data = dat_summaries,
                           bind = TRUE,
                           effect_type = "lnM",
                           verbose = FALSE,
                           SAFE = TRUE,
                           SAFE_boots = 1e6,
                           SAFE_max_secs = 15)
dat_summaries

d_eq <- function(lnM, n1, n2){
  n0 = (2*n1*n2) / (n1 + n2)
  sqrt(2*exp(2*lnM) + (2/n0))
}

# Calculate d_eq:
dat_summaries[, deq := d_eq(lnM = yi_safe, n1 = n1, n2 = n2)]

# Create string labels:
dat_summaries[, estimates := paste0("lnM = ", round(yi_safe, 2), 
                                    ";<br>d<sub>eq</sub> = ", round(deq, 2))]
dat_summaries[, scenario := paste0("m<sub>1</sub>=", round(x1, 1), ", sd<sub>1</sub>=", round(sd1, 1),
                                   "<br>",
                                   "m<sub>2</sub>=", round(x2, 1),  ", sd<sub>2</sub>=", round(sd2, 1))]

# Format distributions to be long and have labels:
distributions.mlt <- melt(distributions,
                            id.vars = c("mean_group_2"))
distributions_final <- merge(distributions.mlt,
                             dat_summaries[, .(mean_group_2, estimates)],
                             by = "mean_group_2")
distributions_final

# Sort:
setorder(distributions_final, mean_group_2)
distributions_final$estimates <- factor(distributions_final$estimates,
                                     levels = unique(distributions_final$estimates))

setorder(dat_summaries, mean_group_2)
dat_summaries$estimates <- factor(dat_summaries$estimates,
                                     levels = unique(distributions_final$estimates))

# distributions_final[, facet_lab := paste("")]
distributions_final$variable <- factor(distributions_final$variable,
                                       c("group_1", "group_2"))

p <- ggplot()+
  geom_density(data = distributions_final, aes(x = value, fill = variable),
               alpha = 0.75)+
  geom_richtext(data = dat_summaries,
            aes(x = 20, y = 0.1, label = scenario), 
            size = 4, label.color = NA, fill = "transparent")+
  facet_wrap(~estimates)+
  ylab("Density")+
  coord_cartesian(ylim = c(0, 0.11))+
  scale_fill_manual(name = NULL,
                    labels = c("group_1" = "Group 1", "group_2" = "Group 2"),
                    values = c("group_1" = "goldenrod", "group_2" = "dodgerblue"))+
  theme_bw()+
  theme(panel.grid = element_blank(),
        strip.background = element_blank(),
        strip.text = ggtext::element_markdown(size = 14),
        legend.position = "bottom")

```

```{r}
#| echo: false
#| eval: true
p
```

The figure shows the the realized means and standard deviations of groups 1 and 2 ($m_{1,2}$, $sd_{1,2}$) after drawing their distributions from the normal distribution.

In the first panel, the means and standard deviations of group 1 and 2 are nearly identical. Yet lnM is -1.94 and **not 0**. Intuitively this is confusing---you'd expect no difference! But lnM doesn't work that way because lnM is the ratio of within to between variance. That's where $d_{eq}$ comes in handy: $d_{eq} = 0.3$. This means that the two groups are ~1/3 of a standard deviation apart---a negligible difference.

In the second panel, $m_2$ is now ~20. lnM is -0.4 and $d_{eq}$ is ~1. This makes sense *because the standard deviation is 5 and $m_2$ is now ~5 away from $m_1$* 

> This can be reported as: "There was *1 standard deviation* between groups"

In the third panel, $m_2$ is 25, lnM is 0.34 and $d_{eq}$ is 2. This also makes sense because the $m_2$ is now 2 standard deviations away from $m_1$. 

> "There were *2 standard deviations* between groups"

As you can see, converting lnM to $d_{eq}$ makes the results intepretable in terms of differences in standard deviation between two groups---just like with SMD. However, unlike SMD, these differences are directionless. It doesn't matter which group is higher or lower: the result will be the same. 

Analyzing lnM, which is a normally distributed and unbounded effect size, allows us to ask evolutionary and ecological questions that have previously been unanswerable with meta-analysis. Converting model estimates and confidence intervals to $d_{eq}$ (using @eq-d-eq-approx) allows us to report these summaries in more or less plain english (well, at least as plain as we can report SMD).

<!-- :::callout-note -->
<!-- ##### Centering lnM to a target absolute SMD -->
<!-- @Nakagawa2026-rb notes that you can also center $d_eq$ on XXXXX. ***HELP*** -->

<!-- When and where to do this??? -->

<!-- ::: -->

### Meta-analytic models with lnM
Because of the distributional properties of lnM and $d_{eq}$, you should fit models directly on lnM instead of on $d_{eq}$. 

However, this can lead to challenges in interpreting the statistical outputs: by default the statistical tests performed by `metafor::rma.mv` compare coefficients to 0: a p-value of 0.01 indicate that a coefficient is significantly different than 0. 

However, for lnM, 0 does not mean the same as it does for other effect sizes: as our simulation above shows, identical means and standard deviations produce an lnM of -1.94. For this reason, we do not recommend reporting p-values for lnM meta-analyses. Instead, convert the model estimates and CIs to $d_{eq}$ (using @eq-d-eq-approx) and interpret them directly.

We'll demonstrate with a dataset provided by `metafor`:

```{r}
data(dat.normand1999)

dat.normand1999 <- eff_size(x1 = m1i, x2 = m1i, 
                           sd1 = sd1i, sd2 = sd2i,
                           n1 = n1i, n2 = n2i,
                           data = dat.normand1999,
                           bind = TRUE,
                           effect_type = "lnM",
                           verbose = FALSE,
                           SAFE = TRUE,
                           SAFE_boots = 1e6,
                           SAFE_max_secs = 15)

m <- rma.mv(yi = yi_safe,
            V = vi_safe,
            data = dat.normand1999,
            random = ~1|study,
            method = "REML",
            test = "t")

m # Comparison to 0
```

The p value and test statistic in the model table above show strong evidence that lnM is significantly different than 0 (`t = -8.8, p < 0.0001`). But 0 is not particularly meaningful.

For that reason, we'll convert the model estimates and CIs to $d_{eq}$ and interpret those in plain english:

```{r}
#| echo: true
#| eval: true

d_eq_approx <- function(lnM){
  sqrt(2) * exp(lnM)
}

broom.mixed::tidy(m) %>% # create a tidy table of model object
  mutate(ci.lb = m$ci.lb,
         ci.ub = m$ci.ub) %>% # Add CIs to table
  mutate(estimate_deq = d_eq_approx(estimate),
         ci.lb.deq = d_eq_approx(ci.lb),
         ci.ub.deq = d_eq_approx(ci.ub)) %>%
  select(term, estimate, ci.lb, ci.ub, estimate_deq, ci.lb.deq, ci.ub.deq) %>%
  knitr::kable()

```

We can then report that the magnitude of differences between groups was ~0.19 standard deviations, with 95% confidence intervals ranging from 0.11 to 0.32 standard deviations. We can use the same approximate benchmarks as we use with [SMD](2.1-SMD-assumptions.qmd#sec-smd-interpretation):

- $d_{eq} \approx |0.2|$ is a weak difference

- $d_{eq} \approx |0.5|$ is a moderate difference

- $d_{eq} \approx |0.8|$ is a strong difference

:::callout-note
#### Testing custom null hypotheses with `anova`
Alternatively, you can use the function `anova` to test a different null value than 0. This is done as follows:

```{r}
anova(m, rhs = -2)
```

This compared the model estimate from our model object `m` to a null value of -2 and returned `F = 0.0001, p = 0.99`---indicating that our estimate is not significantly different than -2. However, choosing what this null value should be is non-trivial. There is no way to solve @eq-d-eq-approx for $d_eq = 0$ (because the result is $-\infty$). For that reason, we don't recommend using `anova` unless there is a good *a priori* justification for a custom null hypothesis value.
:::

### Assumptions {#sec-lnM-assumptions}
**lnM** works with both ratio-scale and interval-scale data and can thus accept negative values. 

However, as with SMD and lnRoM, lnM assumes that the underlying data is more or less normally distributed. We recommend checking for overdispersed data and excluding effect sizes with $CV > 2$, as with SMD and lnRoM. See [here](2.1-SMD-assumptions.qmd) for further discussion of overdispersion. lnM is also sensitive to low sample sizes. @Nakagawa2026-rb recommends conducting sensitivity analyses only including studies with $n_1 + n_2 \geq 40$.

lnM is a newly constructed effect size (as of 2026). Further research is necessary to understand how heteroscedasticity and other numeric properties influence lnM performance.

### Further reading
Please see @Nakagawa2026-rb and the associated [tutorial]((https://itchyshin.github.io/meta-analysis_of_magnitude/)) for more information about lnM.
