---
title: "Differences in magnitude between two groups"
subtitle: "Beyond the mean"
date: today
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**On this Page**"
    output-file: "5.0-differences-in-magnitude.qmd"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
csl: ecology.csl
---


```{r}
#| eval: true
#| echo: false

rm(list = ls())

if(!requireNamespace("metatools")) {
  pak::pak("ejlundgren/metatools")
}
library("metatools")

library("ggplot2")
library("ggtext")
library("data.table")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

```

The effect sizes that we've described in depth in this tutorial have concerned *mean* relationships between two variables or differences in [variability](5.1-variability-between-groups.qmd) between two groups. However, what if you're interested in the magnitude of mean differences, regardless of their direction? These types of questions are common in evolution, where absolute differences between two populations are of interest but directionality of those differences is inconsequential.

Previous attempts to conduct meta-analyses of magnitude have used the absolute value of SMD as an effect size (or 'folded-normal SMD'). However, taking the absolute value of SMD produces effect sizes clustered just above 0 with distributions that violate expectations of normality and can prevent accurate model estimates

**lnM** is a recently developed effect size that describes the absolute difference between two groups and is unbounded and normally distributed [@Nakagawa2026-rb]. 

To calculate lnM, you need the same statistical ingredients as for calculating **SMD** and **lnRoM** effect sizes: sample sizes ($n_1$, $n_2$), means ($m_1$, $m_2$), and standard deviations ($sd_1$, $sd_2$) for two discrete groups.

The math for **lnM** is quite long and built upon classical one-way ANOVA building blocks. Unfold the section below to see the equations:

:::.{callout-note collapse="true"}
#### lnM math
lnM is based on the classical ANOVA building blocks. First we calculate Mean Square Between ($MS_B$) and Mean Square Within ($MS_W$). $MS_B$ estimates the difference between group means. $MS_W$ estimates the variance inside of each group. 

$$
\begin{aligned}
&MS_B = \frac{n_1 n_2}{n_1 + n_2} * (m_1 - m_2)^2\\ 
&MS_W = \frac{(n_1 - 1)sd_1^2 + (n_2 - 1)sd_2^2}{n_1 + n_2 - 2}\\
\end{aligned}
$${#eq-ms}

We then establish within-group ($s_W^2$) and between-group variance ($s_B^2$), standardized by harmonic-mean sample size ($n_0$).

$$
\begin{aligned}
&n_0 = \frac{2n_1n_2}{n_1 + n_2}\\
&s_W^2 = MS_W\\
&s_B^2 = \frac{MS_B - MS_W}{n_0} \\
\end{aligned}
$${#eq-variances}

Finally, with use these ingredients to calculate **lnM** as the ratio of between group standard deviation ($s_B$) and within-group standard deviation ($s_W$):

$$
\begin{aligned}
&lnM = \ln\left(\frac{s_B}{s_W}\right)\\
\end{aligned}
$${#eq-lnM}
:::
### How to calculate lnM {#sec-lnM-how-to}
If the means of group 1 and group 2 are similar, lnM cannot be calculated. For this reason, this effect size should be calculated with SAFE bootstrapping [@Nakagawa2025-do], which simulates hyperparameters based on the summary statistics (i.e., $m_{1,2}$, $sd_{1,2}$) and thus, if repeated enough, can produce estimates of lnM. 

For that reason **lnM** is currently not supported by `metafor::escalc` and has to be calculated in  `metatools::eff_size`. 

Be sure to set `SAFE=TRUE` in the function arguments. The `SAFE_boots` argument indicates how many bootstraps to run (1e6 is default and [appears](https://ejlundgren.github.io/SAFE/) to be a good compromise between accuracy and speed). `SAFE_max_secs` is the number of seconds to run before timing out per effect size calculation. Calculating **lnM** can be time consuming because calculations must be done over and over until a sufficient number of bootstraps is attained.

Let's create a dataset with 2 row for which we want to calculate lnM. Note that the first row of this data has means that are very similar (15 and 15.5) between the two groups:

```{r}
#| eval: true
#| echo: true


if(!requireNamespace("metatools")) {
  pak::pak("ejlundgren/metatools")
}
library("metatools")


dat <- data.table(x1 = c(15, 15), x2 = c(15.5, 20),
                  sd1 = c(2, 2), sd2 = c(2, 2),
                  n1 = c(15, 15), n2 = c(15, 15))
knitr::kable(dat)

out <- eff_size(x1 = x1, x2 = x2, 
         sd1 = sd1, sd2 = sd2,
         n1 = n1, n2 = n2,
         data = dat,
         bind = TRUE,
         effect_type = "lnM",
         verbose = FALSE,
         SAFE = TRUE,
         SAFE_boots = 1e6,
         SAFE_max_secs = 15)

knitr::kable(out)

```

In the resulting table, you'll note that `yi` for the first row is `NA`. This is because the difference between `x1` and `x2` is very small, leading to incalculable values. 

SAFE circumvents this by bootstrapping from the means and standard deviations until a sufficient number of bootstraps is reached (defined by `SAFE_boots`, here set to 1e6). The point estimate and sampling variance produced by SAFE bootstrapping are reported as `yi_safe` and `vi_safe`. Additional information on the performance of the SAFE boostrapping are returned as well (`number_SAFE_iterations`, `number_SAFE_bootstraps_completed`, `SAFE_elapsed_time`, `SAFE_complete`). See function documentation for more information (`?eff_size`).

:::callout-note
Cases where the means between groups are very similar may take a while to successfully run. Increase `SAFE_max_secs` and reduce `SAFE_boots` until your dataset produces finalized SAFE estimates.

To learn more about SAFE see the tutorial [here](https://ejlundgren.github.io/SAFE/) and @Nakagawa2025-do. 
:::

### Interpretation {#sec-lnM-interpretation}
**lnM** is the ratio of between group variation and within-group variation. This between group standard deviation can be thought of as mean-separation---how far apart the two groups are.

lnM can be difficult to interpret on its own. An lnM of 0 corresponds to identical between and within group variation. This should **not** be interpreted as equal group means but whether the between-group dispersion is at least as large as the within-group dispersion. 

While meta-analytic models should be fit on lnM directly, we recommend mapping lnM model estimates to SMD, or $d$, for interpretation to produce $d_{eq}$ or "d-equivalent":

$$
\begin{aligned}
&n_0 = \frac{2 n_1 n_2}{n_1 + n_2}\\
&d_{eq} = \sqrt{2e^{2lnM} + \frac{2}{n_0}}\\
\end{aligned}
$${#eq-d-eq}

Note that when lnM = 0, this translates to $d_{eq}$ of 1.41. What exactly does that mean? This means that within-group variation is large but overall mean differences between groups are 1.41 standard deviations apart. Tricky right?

Let's illustrate lnM in `R` for three comparisons---each representing a plausible comparison of group 1 and 2. We'll draw data from the normal distribution for 3 scenarios: equivalent "true" means, and increasing means for group 2. We'll set the group standard deviations constant at 5, and change the means of group 2 by increments of 5 while holding group 1 constant at $m_1=15$. In order to clearly illustrate how lnM works, we'll set a sample size of 100 for each group. We'll plot the simulated data, lnM estimates, and $d_{eq}$:

```{r}
#| eval: true
#| echo: true
#| output: false
#| code-fold: true
#| fig-width: 10

set.seed(7)

# 
sd <- 5
n <- 500

# Group 1 distribution (fixed):
group_1 <- rnorm(n = n, mean = 15, sd = sd)

# Group 2 distributions:
distributions <- lapply(c(15, 20, 25),
                        function(x) data.table(mean_group_2 = x,
                                               group_2 = rnorm(n = n, mean = x, sd = sd))) |>
  rbindlist()

distributions[, group_1 := rep(group_1, length(unique(distributions$mean)))]

dat_summaries <- distributions[, .(x1 = mean(group_1),
                                   x2 = mean(group_2),
                                   sd1 = sd(group_1),
                                   sd2 = sd(group_2),
                                   n1 = 40,
                                   n2 = 40),
                               by = .(mean_group_2)]
dat_summaries

dat_summaries <- eff_size(x1 = x1, x2 = x2, 
                           sd1 = sd1, sd2 = sd2,
                           n1 = n1, n2 = n2,
                           data = dat_summaries,
                           bind = TRUE,
                           effect_type = "lnM",
                           verbose = FALSE,
                           SAFE = TRUE,
                           SAFE_boots = 1e6,
                           SAFE_max_secs = 15)
dat_summaries

d_eq <- function(lnM, n1, n2){
  n0 = (2*n1*n2) / (n1 + n2)
  sqrt(2*exp(2*lnM) + (2/n0))
}

# Calculate d_eq:
dat_summaries[, deq := d_eq(lnM = yi_safe, n1 = n1, n2 = n2)]

# Create string labels:
dat_summaries[, estimates := paste0("lnM = ", round(yi_safe, 2), 
                                    ";<br>d<sub>eq</sub> = ", round(deq, 2))]
dat_summaries[, scenario := paste0("m<sub>1</sub>=", round(x1, 1), ", sd<sub>1</sub>=", round(sd1, 1),
                                   "<br>",
                                   "m<sub>2</sub>=", round(x2, 1),  ", sd<sub>2</sub>=", round(sd2, 1))]

# Format distributions to be long and have labels:
distributions.mlt <- melt(distributions,
                            id.vars = c("mean_group_2"))
distributions_final <- merge(distributions.mlt,
                             dat_summaries[, .(mean_group_2, estimates)],
                             by = "mean_group_2")
distributions_final

# Sort:
setorder(distributions_final, mean_group_2)
distributions_final$estimates <- factor(distributions_final$estimates,
                                     levels = unique(distributions_final$estimates))

setorder(dat_summaries, mean_group_2)
dat_summaries$estimates <- factor(dat_summaries$estimates,
                                     levels = unique(distributions_final$estimates))

# distributions_final[, facet_lab := paste("")]
distributions_final$variable <- factor(distributions_final$variable,
                                       c("group_1", "group_2"))

p <- ggplot()+
  geom_density(data = distributions_final, aes(x = value, fill = variable),
               alpha = 0.75)+
  geom_richtext(data = dat_summaries,
            aes(x = 20, y = 0.1, label = scenario), 
            size = 4, label.color = NA, fill = "transparent")+
  facet_wrap(~estimates)+
  ylab("Density")+
  coord_cartesian(ylim = c(0, 0.11))+
  scale_fill_manual(name = NULL,
                    labels = c("group_1" = "Group 1", "group_2" = "Group 2"),
                    values = c("group_1" = "goldenrod", "group_2" = "dodgerblue"))+
  theme_bw()+
  theme(panel.grid = element_blank(),
        strip.background = element_blank(),
        strip.text = ggtext::element_markdown(size = 14),
        legend.position = "bottom")

```

```{r}
#| echo: false
#| eval: true
p
```
The figure shows the realized means and standard deviations of groups 1 and 2 ($m_{1,2}$, $sd_{1,2}$) after drawing their distributions from the normal distribution.

In the first panel, the means and standard deviations of group 1 and 2 are nearly identical. Yet lnM is -1.94 and **not 0**. This is obviously confusing---there should be no difference! But lnM doesn't work that way---because lnM is the ratio of within to between variance. That's where $d_{eq}$ comes in handy: $d_{eq} = 0.3$. This means that the two groups are ~1/3 of a standard deviation apart. 

In the second panel, $m_2$ is now ~20. lnM is -0.4 and $d_{eq}$ is ~1. This makes sense *because the standard deviation is 5 and $m_2$ is now 5 away from $m_1$* 

> This can be reported as: "There was *1 standard deviation* between groups"

In the third panel, $m_2$ is 25, lnM is 0.34 and $d_{eq}$ is 2---which again successfully recovered the number of standard deviations between the two groups.

> "There were *2 standard deviations* between groups, regardless of direction"

As you can see, converting lnM to $d_{eq}$ makes the results intepretable in terms of differences in standard deviation between two groups---just like with SMD. However, unlike SMD, these differences are directionless. It doesn't matter which group is higher or lower: the result will be the same. This produces a normally distributed and unbounded effect size distribution which allows us to ask evolutionary and ecological questions that have previously been unanswerable with meta-analysis.

:::callout-important
#### Modelling lnM
Because of the distributional properties of lnM and $d_{eq}$, it is still better to fit models directly on lnM, instead of $d_{eq}$. 

However, this can lead to challenges in interpreting the statistical outputs: the statistical tests performed by `metafor::rma.mv` compare coefficients to 0: a p-value of 0.01 indicate that a coefficient is significantly different than 0. 

However, for lnM, 0 does not mean the same as it does for other effect sizes: as our simulation above shows, identical means and standard deviations produce an lnM of -1.94. 

You can set alternative values using the function `anova` and the argument `rhs` to specify alternative values. 

Let's use $d_{eq}=0$ as our threshold, or 0 standard deviation between groups. We'll use a modified conversion formula between lnM and $d_{eq}$ that assumes large sample size (see equation 22 in @Nakagawa2026-rb). 

$$
\begin{aligned}
&d_{eq} \approx \sqrt{2}e^{lnM} \\
\end{aligned}
$${#eq-deq-large-sample-approx}

Solving for lnM:
$$
\begin{aligned}
&e^{lnM} \approx \frac{d_{eq}}{\sqrt{2}}\\
\end{aligned}
$${#eq-solving-for-lnm}

If $d_{eq} = 0$, this simplifies to...?

Not sure what to do here...

Let's demonstrate with a dataset provided by `metafor`:

```{r}
data(dat.normand1999)
dat.normand1999

dat.normand1999 <- eff_size(x1 = m1i, x2 = m1i, 
                           sd1 = sd1i, sd2 = sd2i,
                           n1 = n1i, n2 = n2i,
                           data = dat.normand1999,
                           bind = TRUE,
                           effect_type = "lnM",
                           verbose = FALSE,
                           SAFE = TRUE,
                           SAFE_boots = 1e6,
                           SAFE_max_secs = 15)
dat.normand1999

m <- rma.mv(yi = yi_safe,
            V = vi_safe,
            data = dat.normand1999,
            random = ~1|study,
            method = "REML",
            test = "t")
m # Comparison to 0

anova(m, rhs = -1.5)

```
:::



:::callout-note
##### Centering lnM to a target absolute SMD
@Nakagawa-rb notes that you can also center $d_eq$ on XXXXX. ***HELP***

When and where to do this???

:::


### Assumptions {#sec-lnM-assumptions}
**lnM** works with both ratio-scale and interval-scale data and can thus accept negative values. 

However, as with SMD and lnRoM, lnM assumes that the underlying data is more or less normally distributed. We recommend checking for overdispersed data and excluding effect sizes with $CV > 2$, as with SMD and lnRoM. See [here](2.1-SMD-assumptions.qmd) for further discussion of overdispersion. lnM is also sensitive to low sample sizes. @Nakagawa2026-rb recommends conducting sensitivity analyses only including studies with $n_1 + n_2 \geq 40$.

lnM is a newly constructed effect size (as of 2026). Further research is necessary to understand how heteroscedasticity and other numeric properties influence lnM performance.


:::callout-important
##### Paired **lnM**
For the paired version of lnM, add `paired = TRUE` to the `eff_size` function call:

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

eff_size(x1 = 15, x2 = 10, 
         sd1 = 1.5, sd2 = 2,
         n = 15, r = 0.8,
         effect_type = "lnM",
         paired = TRUE, # Add this. FALSE is default.
         SAFE = TRUE,
         SAFE_boots = 1e6,
         SAFE_max_secs = 60)


```
:::

### Further reading
Please see @Nakagawa2026-rb and the associated [tutorial]((https://itchyshin.github.io/meta-analysis_of_magnitude/)) for more information about lnM.
