---
title: "Effect sizes"
subtitle: "A soup to nuts tutorial for the meta-analyst"
date: "`r Sys.Date()`"
author:
  - name: Erick J Lundgren
  - name: Eduardo S A Santos
  - name: Malgorzata Lagisz
  - name: Sage Raymond
  - name: Kyle Morrison
  - name: Jonas Trepel
  - name: Eamonn Wooster
  - name: Ayumi Mizuno
  - name: Santiago Ortega
  - name: Iwo Gross
  - name: Wolfgang??
  - name: Shinichi Nakagawa
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**CONTENTS**"
    output-file: "index.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
---

```{r}
#| eval: true
#| echo: false
#| out-width: "70%"

rm(list = ls())
library("metaDigitise")
library("ggplot2")
library("data.table")
library("ggpattern")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

# Source our custom functions:
source("scripts/functions/eff_size.R")
source("scripts/functions/convert_effect_sizes.R")

```

```{r}
#| eval: true
#| echo: false
#| output: false

file.copy("../package_development/metatools_dev/scripts/1_functions/convert_effect_sizes.R",
          "scripts/functions/convert_effect_sizes.R",
          overwrite = T)
file.copy("../package_development/metatools_dev/scripts/1_functions/eff_size.R",
          "scripts/functions/eff_size.R",
          overwrite = T)
file.copy("../package_development/metatools_dev/data/conversion_formulas.csv",
          "data/conversion_formulas.csv",
          overwrite = T)
file.copy("../package_development/metatools_dev/data/effect_size_formulas.csv",
          "data/effect_size_formulas.csv",
          overwrite = T)

```

# Introduction

Meta-analysis has grown to be one of the powerful methods in science. By extracting data from across the literature, a meta-analyst can answer age-old questions at scales previously unimaginable. Moreover, meta-analysis can test for and correct publication bias and can thus not only control for biases in the literature base but provide a bridge between empirical questions (e.g., what is) and metascientific questions.

Meta-analysis can be really hard (and sorry, this tutorial aims to make it easier but doing so also might make it harder because you'll realize how much *you didn't know*). Meta-analytic methods continue to proliferate, with new effect sizes, models, and methods being published constantly. The literature has become a minefield of articles and tutorials that can be bewildering for even the most experienced meta-analyst.

We here present a living tutorial to navigating this methodological tempest, with particular emphasis on extracting and calculating effect sizes from a variety of data formats. This tutorial is maintained by the Collaboration for Open Science and Synthesis in Ecology and Evolution at the University of Alberta, which is led by Professor Shinichi Nakagawa—one of the principle developers of new meta-analytic methods. As such, we hope that this tutorial, by integrating tips and tricks and statistical know-how from across the decades and into the future, can increase the consistency, reliability, and reproducibility of meta-analytic research.

# Voyaging among the ghosts of data past

Meta-analyses are powerful. But they are full of tedium. And dealing with the challenges that we describe in this tutorial (hopefully in plain english) can be overwhelming. 

Yet there is something beautiful and very human about doing a meta-analysis. You encounter the heartwork and hardwork of dozens or hundreds or thousands of fellow human beings. Stories from last year or 50 years ago, from people exploring, observing, experimenting, and trying to understand our universe. 

These voyagers left us breadcrumbs. Sometimes as simple as a mean and a standard deviation. Sometimes as complex as some strange contour plot drawn on a piece of graph paper and scanned in 1975. Extracting their data summaries (or finding their raw datasets) brings their work to life and allows us to continue their quests. 

While you struggle to deal with all the strange and bizarre and difficult and annoying decisions these voyagers sometimes made to communicate their tales, remember to feel compassion and respect and awe for the work that came before us--these efforts to describe the mysteries of our world.

# How to use this tutorial

We begin by providing an overview of the workflow we use in designing a meta-analysis, with particular attention paid to effect sizes, their assumptions, and to testing the influence of those assumptions with sensitivity analyses. We then describe the 3 most common effect size types used in ecology and evolution and their assumptions and alternative formulations. For each of these effect sizes, we then provide a detailed tutorial on how to extract data from both common and bizarre formats.

This tutorial is viewable as a webpage or you can clone the repo and run the .qmd files interactively. This tutorial corresponds with the cheatsheets provided here *XXXXXX.*

This tutorial is based on an `R` workflow and assumes that the users have a basic understanding of `R`.

We begin by giving a brief overview of the steps to conducting a meta-analysis and planning data extraction @sec. We then present an overview of the main effect size types used in meta-analysis and their assumptions @sec. 

We then go into detail about how to extract data for these effect sizes, including the many odd and unusual data formats that you may uncover @sec. We conclude with a discussion of how to convert between effect size types @sec and provide additional code in the Appendix for other unusual cases @sec.

# A living tutorial
This tutorial is a living document. As such, please help us keep it alive and useful. If we missed anything, if we made an error, or overlooked an extraction challenge, please file an Issue on our github page:

https://github.com/ejlundgren/effect_size_tutorial/issues
