---
title: "Simulating non-Gaussian data"
subtitle: "Additional, homeless helpers"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**CONTENTS**"
    output-file: "non_gaussian_responses.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 250px
crossref:
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
---

```{r}
#| eval: true
#| echo: false
#| out-width: "70%"

rm(list = ls())
library("metaDigitise")
library("ggplot2")
library("gt")
library("lubridate")
library("data.table")
library("ggpattern")
library("patchwork")
library("metafor")
library("dplyr")
library("broom")
library("broom.mixed")
library("simplermarkdown")
library("crayon")

# Source our custom functions:
source("scripts/functions/eff_size.R")
source("scripts/functions/convert_effect_sizes.R")

```

## Generating data for non-Gaussian responses

If a study reports mean ± SD for a non-Gaussian response, such as negative binomial (overdispersed count data), then you shouldn't simply reconstruct the underlying data (as in @secs XXX and YYY) with `rnorm`. Instead, you should use the appropriate data generating process. This requires converting the author provided summary statistics (mean and standard deviation) to the flavors required for these non-Gaussian distributions. We will here cover simply how to do this for negative binomial distribution data.

The object `dt` here is a dataset extracted from an article with Mean±SD for two groups in a time series (similar data as in @sec). We want to generate the underlying data for each point, so we can calculate group level mean±SD for the two treatments ("Present" vs "Eradicated"). The data looks like this:

```{r}
#| eval: true
#| echo: false
#| out-width: "70%"

dt <- copy(structure(list(x1 = 1:10, mean = c(1.15246506001662, 0.617012258022042, 
2.92559078415861, 5.44051812652133, 6.57076998232729, 6.95824795972284, 
9.00492379537813, 8.94685646989777, 8.32950404584211, 9.56883252898138
), status = c("Present", "Present", "Present", "Present", "Present", 
"Present", "Present", "Present", "Eradicated", "Eradicated"), 
    group = c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1), sd = c(0.3, 0.3, 
    0.3, 0.3, 0.3, 0.3, 0.3, 0.3, 0.3, 0.3), N = c(10, 10, 10, 
    10, 10, 10, 10, 10, 10, 10), sd = c(0.3, 0.3, 0.3, 0.3, 0.3, 
    0.3, 0.3, 0.3, 0.3, 0.3)), row.names = c(NA, -10L), class = c("data.table", 
"data.frame")))
dt[, sd := sqrt(mean) + .001]

cat(md_table(dt[, .(mean, sd, N, status)]))
```

Note that the standard deviation has a positive relationship with the mean, which is an indicator that this data is derived from a negative binomial distribution.

Now let's convert the author provided `sd` to `size`, which is the parameter that describes the dispersion of the negative binomial family. This is based on the following relationship:

$$\begin{aligned}
Variance = \mu + \frac{\mu^2}{size}
\end{aligned}$$
Solving for `size` (which is the negative binomial flavor of standard deviation):
$$\begin{aligned}
size = \frac{\mu^2}{Variance - \mu}
\end{aligned}$$
Note that Variance must be greater than the mean for the negative binomial distribution. As such, if some of our data has a Variance \< our mean, we'll set Variance equal to the mean.

```{r}
#| eval: true
#| echo: true
#| out-width: "70%"

# First, calculate the size dispersion parameter:
dt[, var := sd^2]
dt[, size := mean^2 / (var - mean)]
dt[var < mean, var := mean]

# Now expand dataset with rnbinom:
dt_expanded <- list()

for(i in 1:nrow(dt)){
  dt_expanded[[i]] <- data.table(y = rnbinom(mu = dt$mean[i],
                                              size = dt$size[i],
                                              n = dt$N[i]),
                                status = dt$status[i])
}

dt_expanded <- rbindlist(dt_expanded)

# Now summarize to calculate our effect size of interest:
final_dt <- dt_expanded[, .(mean = mean(y),
                            sd = sd(y),
                            total_n = .N),
                        by = .(status)]

final_dt
```
